Müller glia cell cycle re-activation by simultaneous cyclin D1 overexpression and p27kip1 knockdown promotes retinal regeneration in mice

Version 1

Abstract

Abstract

Harnessing the regenerative potential of endogenous stem cells to restore lost neurons is a promising strategy for treating neurodegenerative disorders. Müller glia (MG), the primary glial cell type in the retina, exhibit remarkable regenerative abilities in lower vertebrate species, such as zebrafish and amphibians, where injury induces MG to proliferate and differentiate into various retinal neuron types. The regenerative potential of mammalian MG is constrained by their inherent inability to re-enter the cell cycle, likely due to high levels of the cell cycle inhibitor p27Kip1 and low levels of cyclin D1 observed in adult mouse MG. In this study, we found that adeno-associated virus (AAV)-mediated cyclin D1 overexpression and p27Kip1 knockdown exerts a strong synergistic effect on MG proliferation. MG proliferation induced by this treatment was potent but self-limiting, as MG did not undergo uncontrolled proliferation or lead to retinal neoplasia. Single-cell RNA sequencing (scRNA-seq) revealed that cell cycle reactivation leads to immunosuppression and dedifferentiation of MG. Notably, scRNA-seq analysis identified a new cluster of rod-like MG cells expressing both rod and MG genes, which was further validated by RNA in situ hybridization. Cell cycle reactivation also led to de novo genesis of bipolar- and amacrine-like cells from MG. Overall, our findings suggest that AAV- mediated cyclin D1 overexpression and p27Kip1 knockdown stimulate MG proliferation and promote MG reprogramming. This approach may be a promising strategy, especially when combined with other regeneration-promoting factors, to enhance MG-mediated retinal repair.

Introduction

Müller glia (MG) is the last cell type generated by the retinal progenitor cells (RPCs) during development and exhibit a gene expression profile similar to that of late retinal progenitor cells (Jadhav et al., 2009; Roesch et al., 2012). In teleost fish and amphibians, MG respond rapidly to retinal injury, undergoing robust proliferation and regenerating lost retinal neurons from MG-derived progenitor cells (Hamon et al., 2016; Todd and Reh, 2022). In contrast, the proliferative and neurogenic ability of MG in response to injury in mammals is severely limited, failing to mediate retinal self-repair (Karl et al., 2008). Recent investigations have achieved great success in mammalian retinal regeneration by direct conversion of MG to retinal neurons through a single or combination of neurogenic transcription factors (Jorstad et al., 2017; Levi et al., 2021; Todd et al., 2022; Yumi et al., 2015). However, this may lead to a depletion of the MG population and cause further retinal degeneration, as MG are indispensable for retinal function and homeostasis. Other studies have shown that the quiescent state of MG can be overridden by upstream signaling pathways such as Wnt and Hippo (Hamon et al., 2019; Rueda et al., 2019; Yao et al., 2016). Activation of Wnt signaling by forced expression of β-catenin in adult mouse MG promoted spontaneous cell cycle re-entry in uninjured retinas (Yao et al., 2016). Moreover, it was demonstrated that bypassing the Hippo pathway in mouse MG led to spontaneous re-entry into the cell cycle and reprogramming into a progenitor cell-like state (Hamon et al., 2019; Rueda et al., 2019). These findings suggest that the re-entry of MG into the cell cycle, which is the first step of MG-mediated retinal regeneration in zebrafish, could be unlocked in mammalian retinas.

The cell cycle of MG is mainly regulated by cyclins and cyclin-dependent kinases (CDKs). Cyclins bind to CDKs to form cyclin-CDK complexes to promote cell cycle progression. During retinal development, the expression of D-type cyclins (cyclin D1, D2, and D3) is tightly regulated (Barton and Levine, 2008; Dyer and Cepko, 2001; Trimarchi et al., 2008). Among these, cyclin D1, encoded by the Ccnd1 gene, is the predominant D-type cyclin in the developing retina and is highly expressed in the RPCs but absent in differentiated cells (Barton and Levine, 2008; Trimarchi et al., 2008). Mice lacking Ccnd1 have small eyes and hypocellular retinas due to reduced RPC proliferation (Fantl et al., 1995; Sicinski et al., 1995), which cannot be compensated by Ccnd2 and Ccnd3 (Das et al., 2012, 2009). Negative regulators of the cell cycle are CDK inhibitors (CDKIs), which include the INK4 (p16INK4a, p15INK4b, p18INK4c, and p19INK4d) and CIP/KIP families (p21Cip1, p27Kip1, and p57Kip2) (Reynisdóttir et al., 1995). The CDKIs inhibit cell cycle progression by binding to and inactivating the cyclin-CDK complexes (Besson et al., 2008), and they regulate the proliferation in distinct retinal progenitor cells (Dyer and Cepko, 2000a, 2001; Levine et al., 2000). p27Kip1 inhibits the cyclin D-CDK complex to enter of the S phase. Following acute retinal damage in mice, a very small number of MG re-enter the cell cycle, co-incident with the downregulation of p27Kip1 or upregulation of cyclin D1 (Dyer and Cepko, 2000b; Hamon et al., 2019; Rueda et al., 2019; Yao et al., 2016). However, this process is transient, as cyclin D1 expression rapidly returns to the basal level (Dyer and Cepko, 2000; Hamon et al., 2019a; Rueda et al., 2019).

In this study, we propose that targeting the two key regulators of the cell cycle, cyclin D1 and p27kip1, can effectively induce MG cell cycle re-entry. Our findings demonstrate that simultaneously reducing p27Kip1 and increasing cyclin D1 in the MG using a single AAV vector has a strong synergistic effect on promoting MG proliferation in uninjured adult mouse retinas. MG proliferation induced by this treatment is robust and self-limiting, as MG undergo a single round of cell division rather than unlimited proliferations. Through single-cell RNA sequencing (scRNA-seq), we observed that cell cycle reactivation leads to immunosuppression and dedifferentiation of the MG. Interestingly, scRNA-seq analysis also showed that a new cluster of rod-like MG cells that express both rod and MG genes was induced by the treatment.

RNA in situ analysis further confirmed that MG cells in the outer nuclear layer (ONL) express high levels of rod gene Rho and Gnat1 while maintaining the expression of MG gene Glul. Additionally, a few EdU+ MG daughter cells in the inner nuclear layer (INL) express the bipolar cell marker Otx2 or the amacrine cell marker HuC/D, suggesting de novo neurogenesis from MG. These results indicate that cell cycle reactivation also promotes MG reprogramming. Importantly, MG cell cycle reactivation does not disrupt retinal structure or impair retinal function, as the treatment did not deplete the MG from the retina. In summary, our results showed that MG cell cycle reactivation by downregulating p27Kip1 and upregulating cyclin D1 stimulates MG proliferation and promotes cell fate reprogramming, and it is possible to combine this approach with other factors that promote regeneration in order to enhance retinal repair mediated by MG.

Results

Simultaneous p27Kip1 downregulation and cyclin D1 overexpression drive robust MG proliferation in the uninjured mouse retina

To test the hypothesis that adult mouse MG are kept in a quiescent state by high levels of p27Kip1 and low levels of cyclin D1, we examined whether spontaneous MG proliferation can be activated by directly changing the levels of these two downstream cell cycle regulators. In this study, we used adeno-associated virus (AAV) serotype 7m8 vector and a promoter sequence cloned from the human glial fibrillary acidic protein (GFAP) gene to drive MG- specific gene expression (Fig. 1A, Supplementary Fig. 1) (Lee et al., 2008). AAV7m8 vectors were injected intravitreally into the eyes of C57BL/6 mice on postnatal day 6 (P6), when MG finish proliferation and start differentiation, and intraperitoneal injections of 5-ethynyl-2’- deoxyurdine (EdU) were given for five consecutive days to label the MG that had entered the S phase of the cell cycle (Fig. 1B). In the control retinas infected with AAV7m8-GFAP-GFP- nontarget (NT) short hairpin RNA (shRNA) virus, all MG, which were identified by Sox9 expression aligning in the inner nuclear layer (INL), were negative of EdU labeling (Fig. 1C). When the retina was infected by AAV7m8-GFAP-mCherry-p27Kip1 shRNA, which expresses a highly efficient p27Kip1 shRNA, a small number of MG cells re-entered the cell cycle (Fig. 1D, G, Supplementary Fig. 2); however, these proliferated MG accounted for less than 1% of the total MG population. Similarly, overexpressing cyclin D1 alone through AAV7m8-GFAP- cyclin D1 infection stimulated a subset of MG cells to proliferate, resulting in a three-fold increase in the number of EdU+ MG cells compared to p27Kip1 knockdown (Fig. 1E, G, Supplementary Fig. 3). Finally, the AAV7m8-GFAP-cyclin D1-p27Kip1 shRNA vector, which simultaneously overexpressed cyclin D1 and suppressed p27Kip1, had the most significant impact on MG proliferation, with a five-fold increase in EdU+Sox9+ cells compared to cyclin D1 overexpression alone (Fig. 1F, G, whole retinal cross-section shown in Supplementary Fig. 4). These findings suggest that both low levels of p27Kip1 and high levels of cyclin D1 are required for postmitotic MG to re-enter the cell cycle. Henceforth, we refer to this AAV vector that enables robust MG proliferation through concurrent p27Kip1 knockdown and cyclin D1 overexpression as the cell cycle activator (CCA).

Simultaneous p27Kip1 downregulation and cyclin D1 overexpression drive robust MG proliferation in the uninjured mouse retina

(A) Schematic of AAV vectors used in this study. (B) Experimental design. (C-F) Analysis of EdU incorporation. Uninjured mouse eyes were injected with (C) AAV-GFAP-mCherry-non target (NT) shRNA (control), (D) AAV-GFAP-mCherry-p27 shRNA, (E) AAV-GFAP-cyclin D1, and (F) AAV-GFAP-cyclin D1-p27 shRNA (CCA) at P6 and harvested at P18 after 5-day EdU intraperitoneal injection. Retinal sections were co-labeled for MG marker Sox9. (G) Quantification of EdU+ Sox9+ cells. AAV-GFAP-mCherry-NT shRNA control virus (n=8), AAV-GFAP-mCherry-p27shRNA (n=11), AAV-GFAP-cyclin D1 (n=8), and CCA (n=14). (H) Quantification of the percentages of EdU+GFP+ and EdU-GFP+ cells in the area with efficient virus infection in the Glast-CreERT2; Sun1:GFP mouse retina injected with the control or CCA vector. (I) Quantification of the total GFP+ MG in the CreERT2; Sun1:GFP retina infected by CCA. Retinal area (250 µm in width) with the most efficient virus infection in the middle retina region was selected for quantification. (J) Quantification of EdU+ Sox9+ cells in the young (CCA injection at P6, n=13), adult (CCA injection at P28, n=17), and aged retinas (CCA injection at P255-P347, n=7). ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer. Data are presented as mean ± SEM. * P<0.05; *** P<0.001; ns=not significant by one-way ANOVA analysis with Tukey’s post-hoc test (G, J) and two-tail unpaired student t-test (I).

A transgenic mouse line Glast-CreERT2; Sun1:GFP, in which MG nuclei were labeled by nuclear membrane-bound GFP (Supplementary Fig. 5), was used to quantify the percentage of MG that re-entered the cell cycle. In the retinal area where virus infection rate was the highest, approximately 45% of Sun1:GFP+ MG were EdU positive (Fig. 1H), and the total number of MG increased by about 50% (Fig. 1I), indicating that nearly half of the MG cells re-entered the cell cycle. Previous research has shown that the ability of retina regeneration by various stimuli declines with the age of the mice (Löffler et al., 2015; Ueki et al., 2015). We compared the efficiency of CCA in driving MG proliferation in young (P28) and aged (>8 months) mouse retinas to that in P6 pups. Remarkably, MG proliferation induced by CCA remained robust in young adult and aged mice, with more than 200 EdU+Sox9+ cells per section (Fig. 1J). These findings suggest that CCA efficiently drives MG proliferation in both young and aged animals.

MG proliferation driven by CCA is self-limiting

Concerned that p27Kip1 suppression and cyclin D1 overexpression may lead to uncontrolled cell proliferation and retinal tumorigenesis, we analysed the proliferative capacity of MG driven by CCA. Firstly, we examined the duration of MG proliferation after CCA treatment by a time-course EdU incorporation assay. On various days after the CCA injection, EdU was administrated to label the cells that were undergoing proliferation (Fig. 2A). The result revealed that MG proliferation started as quickly as the third day after CCA injection, reached its peak around the fifth day, followed by a gradual decrease (Fig. 2A). By two weeks post CCA injection, only a few MG were observed to have re-entered the cell cycle. Two months post- CCA injection, MG proliferation had mostly ceased (Fig. 2A). These findings suggest that MG proliferation was largely completed within 2 weeks post-CCA treatment, probably due to the dilution of AAV episomal genome copies in the dividing cells. In addition, an EdU/BrdU double-labeling assay was performed to examine whether MG undergoes one or multiple cell divisions after CCA treatment. A single injection of EdU was given at 7 days post CCA treatment, followed by BrdU injection 24 hours later (Fig. 2B). Retinas were collected 2 days after BrdU injection to evaluate if any MG continuously entered the S phase of the cell cycle. While there were a number of EdU+ cells and BrdU+ cells, few cells were co-labeled with EdU and BrdU (Fig. 2B,C), indicating that MG undergoes only one cell division subsequent to CCA treatment.

MG proliferation driven by CCA is self-limiting

(A) Time-course analysis of MG proliferation after CCA injection. EdU was administered for two consecutive days, starting from different days after the CCA injection, and samples were harvested for analysis one day after the second EdU injection. Data are presented as mean ± SEM, n≥4. (B) Analysis of EdU and BrdU labeled cells. (C) Quantification of the number of EdU+BrdU-, EdU-BrdU+ and EdU+ BrdU+ cells. Data are presented as mean ± SEM. n≥4. (D) Representative retinal sections of Glast-CreERT2; Sun1:GFP mice at one week, two weeks, and four months post CCA injection. (E) Quantification of MG daughter cell (EdU+GFP+) distribution in each retinal layer. ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer. Data are presented as mean ± SEM, n=3.

Next, we examined the distribution of the proliferated MG in the retina over time. As the labelling efficiencies of MG by Tamoxifen-induced reporter expression and by EdU varied among experiments and individual animals, we used the percentages of EdU+GFP+ MG in each retinal layer to assess the distribution of the proliferated MG. During retina development, proliferating retinal progenitor cells migrate in an apical-basal manner, whereas post-mitotic cells migrate basally to reach their final laminar position (Lee and Norden, 2013). The proliferating MG undergo similar interkinetic migration towards the apical surface of the retina, as majority of the EdU+GFP+ MG nuclei were observed in the ONL and OPL (∼53% and ∼22%, respectively) at one week after CCA treatment, which is around the peak time of MG proliferation (Fig. 2D,E). To further explore whether the ONL MG daughter cells eventually return to the INL over an extended period, we examined the distribution of EdU+ MG cells at three weeks and four months post-CCA treatment. We found that while the percentage of INL EdU+ MG increased slightly over time, there was still about half of EdU+ MG remained in the ONL and in the OPL (Fig. 2D,E). Across the three time points examined, the absence of EdU+GFP+ MG in the IPL and GCL suggests that migration of MG daughter cells to the GCL does not occur, at least without injury (Fig. 2D,E). These findings led us to investigate whether the long-term localization of MG daughter cells in the ONL and OPL is associated with any change in their cell fate.

Cell cycle re-activation drives downregulation of IFN pathway and MG dedifferentiation

To unbiasedly assess the cell fate of MG and their daughter cells, we performed single-cell RNA-Seq (scRNA-Seq) analysis on the mouse retinas that received CCA treatment. AAV7m8- GFAP-cyclin D1- p27Kip1 shRNA and AAV7m8-GFAP-GFP-NT shRNA (9:1 mixed) were co- injected to stimulate MG proliferation in 4-week-old Glast-CreERT2;tdTomato mice (Fig. 3A), and 10% AAV7m8-GFAP-GFP-LacZsh was added to serve as an indicator of infection success. The control group was injected with AAV7m8-GFAP-GFP-NT shRNA to account for any non- specific effect caused by virus injection and/or shRNA expression. Three weeks after virus injection, the retinas with successful viral transduction were collected, and tdTomato+ cells were isolated by fluorescence-activated cell sorting (FACS), followed by scRNA-seq (Fig. 3A). As previous studies demonstrated that N-methyl-D-aspartate (NMDA)-induced retinal damage and histone deacetylase inhibitor Trichostatin A (TSA) improve MG proliferation and reprogramming (Hoang et al., 2020; Jorstad et al., 2017; Rueda et al., 2019) , a group of CCA- treated mice also received NMDA on day 7 and TSA on day 9 post-CCA injection (CCA+NMDA+TSA, referred to as CCANT) to enhance the reprogramming effect, if any, of CCA.

scRNA-seq analysis of MG at three weeks post CCA treatment

(A) Schematic illustration of scRNA-seq experiment. (B) UMAP plot of scRNA-seq data for MG treated with CCA, CCANT, and control virus, with clusters identified by the known marker gene expression. (C) Expression of retinal cell markers in different cell clusters. (D) Split UMAP plots of the control, CCA, and CCANT groups. (E) The proportion of cell clusters in the control, CCA, and CCANT groups. (F) Heatmap of top DEG genes between cell clusters. Cells are shown in columns, and genes are in rows. Color scale denotes Z score of the normalised gene expression levels. (G) The violin diagram shows the expression of IFN pathway, MG and Rod genes in different cell clusters. (H) Feature plots of cell cycle regulator (Mik67, Mcm5), glial (Glul, Kcnj10), gliosis (Gfap), cell proliferation (Btg2), and rod (Rho and Nrl) normalised gene expression in different cell clusters. CCA, cell cycle activator; CCANT, cell cycle activator + NMDA + TSA.

After quality control, 3,758 cells were profiled in the control group, 3,890 cells in the CCA group, and 3,278 cells in the CCANT group (Supplementary Figures 6 and 7). Clustering analysis separated the cells into six different clusters (Fig. 3B), which are quiescent MG, reactivated MG, MG in G2/M phase, MG in S phase, rod-like MG, and rods, as annotated by the known retinal cell type markers (Fig. 3C). In the control group, the vast majority of MG (>90%) remained as quiescent MG which express high levels of MG genes such as Glul and Kcnj10 (Fig. 3C-H). Cell cycle state analysis showed that a small percentage of cells are in the G2/M or S phase in the CCA group, suggesting that there was still a small number of proliferating MG at three weeks after CCA treatment (Fig. 3H, Supplementary Figure 7D,E). In the CCA group, the majority of MG (∼70%) formed a separate cluster, which we refer to as reactivated MG. In this cluster of cells, we observed downregulation of MG genes (such as Glul and Kcnj10) and upregulation of reactive gliosis gene Gfap and cell cycle inhibitor Btg2 compared with the quiescent MG cluster (Fig. 3H). Interestingly, the top differentially expressed genes (DEGs) between reactivated MG and quiescent MG are the interferon (IFN) pathway genes, including Stat1, Stat2, Irgp1, Irgm1 and Igtp, which were downregulated in the reactivated MG (Fig. 2F-H, Supplementary Fig. 8), while Stat3 was not downregulated concurrently with Stat1 and Stat2 (Supplementary Fig. 8). Here, downregulation of the IFN pathway in the reactivated MG may be a direct response to cyclin D1 overexpression and/or p27Kip1 knockdown. A small percentage of MG (∼8%) of reactivated MG was also present in the control virus-injected group, which may be due to retinal injury caused by virus injection and inflammatory response caused by the AAV virus (Fig. 3D, E). The cell clusters of the CCANT group largely overlapped with the CCA group, with CCANT having fewer proliferating cells and more reactivated MG.

CCA upregulates rod gene expression in a subset of MG

Interestingly, the rods and rod-like MG were two unexpected clusters that appeared in the scRNA-seq analysis. Upon close examination of the retina of Glast-CreERT2; R26-tdTomato mice without Tamoxifen induction, we found that very few rods (14 Tdt+ rods in 128 whole retinal sections of eight mouse retina samples screened) were mislabelled by leaky Tdtomato expression in a Cre-independent manner (Supplementary Figure 9). The small rod clusters are likely the true rods mislabelled by Tdt, which is presented in all three groups with a similar proportion. The rod-like MG cluster resides between the reactivated MG cluster and the rod cluster, and it is significantly enriched in the CCA and CCANT groups. The rod-like MG expressed most rod genes, including Rho, Gnat1, Rcvrn, Crx, and Nrl (Fig. 3G, H, Supplementary Fig. 10), although the expression levels of the rod-specific genes were lower than those of natural rods. However, not all the rod genes are evenly upregulated, as the upregulation of transcription factor Otx2 that is crucial for rod cell fate specification and the genes related to neuron projection and photoreceptor cell cilium such as Syt1 and Cacnb2 were less significant (Supplementary Figure 10). In addition, the rod-like cells expressed a high level of MG genes, while natural rods did not (Fig. 3G, H). When comparing the rod-like MG cluster with the reactivated MG cluster, gene ontology (GO) analysis showed significant gene enrichment in the pathways related to visual system development, neuron differentiation, visual perception, and neuron projection (Supplementary Fig. 10). And GO was significantly enriched in the genes associated with neuron projection and neuron differentiation in the rod cluster as compared to the rod-like cluster (Supplementary Fig. 10), suggesting that rod-like MG were not fully differentiated rods. When comparing the CCA- and CCANT-treated groups, NMDA and TSA increased the percentages of reactivated MG and rod-like MG, but they did not induce the fully differentiated rods or any other new cell clusters. In the following study, we focused solely on the CCA treatment.

Rod-like MG in the ONL expresses both rod and MG genes

To exclude the possibility that the rod-like MG are fused MG with rod cell debris during retina isolation for scRNA-seq analysis, we examined the expression of rod gene expression in the rod-like MG on the retinal sections of the CCA-treated Glast-CreERT2;Sun1:GFP mice. As the immunoreactivity for rod proteins such as RHO and GNAT1 is present in rod outer segment (ROS), which are densely packed and distal from rod nuclei, we performed RNA in situ hybridization on retinal sections using the RNAscope assay, which allowed us to analyse target RNA abundance in rod nuclei with the preservation of the spatial information. We examined the mRNA levels of the Rho and Gnat1 genes at three weeks post CCA treatment, which is the same harvest time point as in the scRNA-seq data. The signals of both probes are specific as they are only detected in the ONL layer (Fig. 4A,B). In the control retinas, the mRNA of the two rod genes was abundantly present in the rods in the ONL (Fig. 4A,C), while there was minimal Gnat1 and Rho mRNA in the MG in the INL (Fig. 4A,E). In the CCA-treated retina, the MG in the ONL showed clear signals of both Gnat1 and Rho mRNA (Fig. 4B,D), while the INL MG maintained very low levels of Gnat1 and Rho mRNA (Fig. 4B,F). We counted the number of Gnat1 fluorescent dots, which served as a reliable indicator of Gnat1 mRNA levels. The result showed a significant upregulation of Gnat1 mRNA levels in the MG-derived cells in the ONL and that its level was comparable to the natural rods (Fig.4G). The MG in the OPL had intermediate levels of Gnat1 mRNA, while the MG in the INL is similar to the untreated MG controls (Fig. 4G). The Rho mRNA levels were quantified by measuring the pixel intensity due to the fusion of numerous Rho fluorescent dots, which showed a similar trend of Rho mRNA expression as that of Gnat1 (Fig. 4H). The RNAscope result demonstrated that the MG- derived cells in the ONL express high level of rod genes, those in the OPL express intermediate level of rod genes, while all MG in the INL retained the no/low rod gene expression, suggesting that the expression level of rod genes in the MG daughter cells is associated their position within the retinal layers.

Gnat1 and Rho mRNA are expressed in the ONL and OPL MG

(A-B) Gnat1 and Rho mRNA in situ hybridization in the control (A) and the Glast- CreERT2;Sun1:GFP mouse retinas harvested at three weeks post CCA injection (B). (C-F) Magnified views of the highlighted regions in (A) and (B). (G) The number of Gnat1 mRNA dots per GFP+ cell and (H) the average pixel level of Rho mRNA per GFP+ cell. n=3 mice, data are presented as mean ± SEM. ns=not significant, *P < 0.05; **P < 0.01; ***P < 0.001; ****P < 0.0001, by one-way ANOVA with Tukey’s post hoc test (G, H).

We further examined the expression level of the MG marker gene Glul, which encodes Glutamate synthetase, in the MG cells located in different retinal layers (Fig. 5A-D). After CCA treatment, the INL MG continue to express Glul, although the level is decreased (Fig. 5E), which is consistent with the downregulation of MG gene in the majority of the MG by scRNA-seq. The MG-derived cells in the ONL expressed lower level of Glul, suggesting that they retain the MG identity as well. The RNAscope results align with the scRNA-seq analysis, indicating that the MG-derived cells in the ONL express both rod and MG genes at three weeks post CCA treatment.

Glul mRNA level is decreased in the ONL and OPL MG

(A-B) Glul mRNA in situ hybridization. (C-D) Magnified views of the highlighted regions in (A) and (B). (E) The average pixel level of Glul mRNA per GFP+ cell. n=3 mice, data are presented as mean ± SEM. ns=not significant, *P < 0.05; **P < 0.01; ***P < 0.001; ****P < 0.0001, by one-way ANOVA with Tukey’s post hoc test (E).

To find out whether the ONL MG could differentiate into mature rods if given an extended period of time, we examined the morphology of the EdU+ MG in the ONL. Despite the extensive search, we did not identify any EdU+ Tdt+ cells in the ONL with a ROS, which is an important functional structure for rod phototransduction (Fig. 6A,B). Mature mouse rods have a unique nuclear architecture, where heterochromatin localizes in a single large chromocenter surrounded by euchromatin (Solovei et al., 2009). After CC treatment, the majority of the ONL MG nuclei do not resemble the rod nuclei (Supplementary Figure 11A- D). In addition, we labeled the MG sparsely with a low dose of Tamoxifen before CCA treatment and found that most of the ONL MG daughter cells retained the apical processes (Supplementary Figure 11E). We examined the mRNA expression of rod and MG genes again at four months after CCA treatment. Rho, Gnat1, and Glul mRNA expression do not display significant change at four months post-CCA treatment (Supplementary Fig. 12). Therefore, we conclude that the ONL MG do not develop to mature rods despite the continuous expression of rod gene and that they retained MG properties.

CCA induces de novo genesis of Otx2+ BP-like and HuC/D+ AC-like cells from MG

(A-B) Representative retinal sections of Glast-CreERT2; tdTomato mice at four months post injection of CCA. Sections were immunostained with EdU and Otx2. (C-D) Magnified views of the highlighted regions in (B). (E) Quantification of tdT+EdU+Otx2+/ tdT+EdU+%. (F-G) Representative retinal sections immunostained with EdU and HuC/D. (H-I) Magnified views of the highlighted regions in (F) and (G). (J) Quantification of tdT+EdU+HuC/D+/tdT+EdU+%. n=3 mice, data are presented as mean ± SEM. *p<0.05; **p<0.01, by unpaired two-tailed Student’s t-test (E, J).

Few bipolar and amacrine-like cells are regenerated from MG

Bipolar (BP) and amacrine cells (AC) are two retinal neuron types that have been shown to be regenerated from MG (Hoang et al., 2020; Jorstad et al., 2020, 2017). Although the scRNA- seq data did not reveal other retinal neuron-like clusters derived from MG after CCA treatment, we performed immunostaining of the bipolar marker Otx2 or the ganglion/amacrine maker HuC/D to examine whether there are rare MG-derived cells expressing either marker. In order to specifically identify de novo neurogenesis from MG and to exclude any mislabelled neurons in the reporter mouse line, we focused our analysis on the EdU+Tdt+ MG in the Glast-CreERT2; R26-tdTomato mice. At four months post-CCA treatment, we observed less than 1% of the total EdU+ Tdt+ cells expressing Otx2, whereas these cells were not present at three weeks post-CCA treatment (Fig. 7A-E). We also found rare HuC/D+EdU+Tdt+ cells in the lower INL where the HuC/D+ amacrine cells naturally reside (Fig. 7F-J). Whether these cells express other BP or AC markers and whether they are functional interneurons that connect with the retinal circuitry needs further investigation. Nonetheless, our findings suggest that cell cycle re-activation alone can increase the plasticity of MG daughter cell fate, which allows some of them to become BP or AC-like neurons.

CCA does not lead to retinal neoplastic transformation

(A-C) Optomotor and electroretinography (ERG) tests were performed on wild type mice at 1 year after CCA injection (n≥4). Visual acuity by the optomotor test (A), and b-wave amplitudes of the scotopic ERG under different light intensity in (B) and photopic ERG under 30 cd × s/m2 in (C), respectively. Data are presented as mean ± SEM. ns=not significant by two-tail unpaired Student’s t-test (A-C). (D-E) Sox9 immunofluorescence in untreated control mice (D) and WT mice one year after CCA injection (E). (F-G) Zoom-in images of the indicated area in (D) and (E). (H) Quantification of the number of Sox9+ cells in retinas at two weeks or one year after CCA injection and in age-corresponding WT control retinas. (I) Quantification of the number of Sox9+ cells in each retinal layer. Data are presented as mean ± SEM. ns=not significant, **p ≤0.01, ***p ≤0.001 by one-way ANOVA with Tukey’s post hoc test (H) and two-tail unpaired Student’s t-test (I).

CCA does not cause retinal neoplasia or functional deficit

To assess the long-term effect of CCA treatment, a cohort of C57BL/6 mice were observed for a year following CCA injection (Fig. 7). Visual function assessments showed no significant differences in visual acuity and electroretinography (ERG) between CCA-treated eyes and uninjected control eyes one year after CCA treatment (Fig. 7A-C). Upon examining all the harvested retinas from the CCA-treated group, no retinal neoplasia was observed. The retinal structure remained intact, with no malignancy or disruptions in the retinal layers or stratification (Fig. 7D,G). These results indicate that CCA treatment did not have a detrimental impact on retinal structure or function in mice, nor did it induce retinal tumours.

We used the MG marker Sox9 to examine the MG population in the retina sections harvested at a year post CCA treatment. In the control retina, Sox9+ MG cells were aligned in the INL (Fig. 7D-E). In the CCA-treated retinas, there was a significant expansion in the number of Sox9+ MG cells, distributed across the ONL, OPL, and INL (Fig. 7F-G). The large population of Sox9+ MG remained to support the retinal homeostasis, which explains the relatively unaffected retinal structure and function observed on the CCA-treated eyes. We further made comparison of the numbers of Sox9+ cells after one year of CCA treatment and after 2 weeks of CCA treatment. The total numbers of Sox+ cells were similar between these two time points (Fig. 7H), indicating no significant MG loss after MG proliferation. To directly assess whether CCA treatment may lead to the cell death of the MG, especially the MG displaced in the ONL and OPL, we examined the CCA-treated retinas with TUNEL assay and did not observe many TUNEL+ cells (Supplement Figure 13). Since we could only examine cell death at single time points, we cannot completely exclude the possibility that there were small numbers of MG-derived cells ding without being detected. The fact that the numbers of MG in the ONL and OPL did not decrease in the retinas harvested at one-year post-CCA treatment suggests that these MG-derived cells persist without dying or losing the MG identity (Fig. 7I).

Discussion

MG proliferation is a key step towards MG-mediated regeneration in the retina. Previously, it was shown that activation of the canonical Wnt pathway by AAV-mediated overexpression of β-catenin stimulates MG proliferation through the Lin28/let7 pathway (Yao et al., 2016), and cyclin D1 is a direct target gene of both Wnt signaling and Lin28/let7 (Li et al., 2012; Shtutman et al., 1999; Tetsu and McCormick, 1999). Suppression of the Hippo pathway by transgenic expression of YAP5SA induces MG proliferation, accompanied by an increase in cyclin D1 expression (Hamon et al., 2019; Rueda et al., 2019). However, the Hippo and Wnt pathways have downstream genes besides cyclin D1 and serve different cellular functions. It is not clear whether cyclin D1 activation is necessary or sufficient to drive MG proliferation in these contexts. In this study, we demonstrated that cyclin D1 overexpression alone can promote MG proliferation and that the combination of cyclin D1 overexpression and p27Kip1 KD is the most potent to drive MG proliferation in mouse retina without injury stimulus.

The strategy to stimulate the MG cell cycle for the purpose of retinal regeneration must be carefully weighed against the concern of potential tumorigenesis. p27Kip1 and cyclin D1 are gatekeepers of cell cycle regulation, and any abnormalities in their expression may impact cell division, potentially leading to the development of tumours. Previous studies have provided evidence that mice lacking p27Kip1 are approximately 30% larger and generally do not exhibit malignancies, except for pituitary tumours that develop spontaneously (Fero et al., 1996). Cyclin D1 is frequently overexpressed in a significant fraction of human cancers, such as breast and respiratory tract tumours (Callender et al., 1994; Zukerberg et al., 1995). The concern regarding retinal tumour formation due to CCA treatment may be mitigated, as the in vivo gene transfer was performed using AAV vectors. The advantage of using AAV vector is that AAV episomal genome becomes diluted after cell division, reducing the risk of tumor formation due to continuous cyclin D1 overexpression and p27Kip1 KD. Consistent with this, the rate of MG proliferation was observed to decrease gradually over time, and the majority of MG cells did not undergo multiple cell divisions (Fig. 2). Additionally, the retinal structure was normal without neoplasia (Fig. 6,7). These results suggest that CCA is a potent and viable treatment to stimulate MG proliferation and may be a promising approach to be used in combination with other neurogenic factors to promote MG-mediated retinal regeneration.

Interestingly, we found that the IFN pathway genes are enriched in the top downregulated genes in the reactivated MG cluster. The IFN pathway, mediated by STAT1 and STAT2, is an important component of the innate immune system against viral infection and is also known as antitumor cytokine signaling which facilitates immunosurveillance in tumor cells (Durbin et al., 1996; Kaplan et al., 1998). This pathway has also been implicated in suppressing regeneration in mice (Jorstad et al., 2020; Todd et al., 2020). STAT activation was shown to reduce the ASCL1-induced retinal regeneration (Jorstad et al., 2020), and decreased IFN signaling could be induced by microglia depletion, which improves MG neurogenic capacity (Todd et al., 2020). In this study, we showed that the IFN pathway genes were downregulated in the MG after CCA treatment. This would be a direct response to the upregulation of cyclin D1 and/or downregulation of p27kip1 in the MG cell-autonomously, or it could indirectly result from the microglial responses elicited by the treatment. Investigating whether and how the downregulation of IFN signaling plays a role in CCA-induced retinal regeneration would be interesting questions to explore further.

The upregulation of rod genes in the MG residing in the ONL is an unexpected outcome of MG cell cycle activation by CCA, as this was not reported in the other studies focusing on MG proliferation (Hamon et al., 2019; Rueda et al., 2019; Yao et al., 2016). In those studies, a very small subset of proliferative MG differentiated into BP and/or AC-like, but no photoreceptor genesis was reported. In our study, a significant subset of MG expressed the mRNA of Rho and Gnat1 by scRNA-seq and RNAscope. However, the rod-like MG is unlikely a result of MG-to-rod cell fate reprogramming, as we did not see many cells overexpressing neurogenic genes, such as Ascl1 and Dll3, or rod precursor gene prdm1 (Supplementary Figure 14A). scRNA-seq analysis at earlier time points after CCA treatment would be helpful to further exclude this possibility. How does CCA activate rod gene expression if it is not through cell fate reprogramming? One hypothesis is that cyclin D1 or p27kip1 regulates rod gene expression in addition to their well-characterized functions in cell cycle regulation. It was shown that cyclin D1 possesses transcriptional functions by recruiting CREB-binding protein histone acetyltransferase in developing mouse retinas (Bienvenu et al., 2010). Cyclin D has been shown to activate retinoblastoma protein (Rb) by mono-phosphorylation (Narasimha et al., 2014), and Rb in turn promotes rod maturation, which is independent of its function for RPC cell cycle exit (Zhang et al., 2004). In the Rb1−/− retinal explants, many genes expressed in rods, such as Nrl and Nr2e3, were significantly downregulated (Zhang et al., 2004). Additionally, p27xic1, the Xenopus homolog of mammalian p27kip1, has been shown to promote glial cell fate independent of its CDK binding activity (Ohnuma et al., 1999). Therefore, besides cell cycle activation, cyclin D1 overexpression and/or p27kip1 KD may promote rod gene expression directly or indirectly, which may explain the rod-like MG population observed following CCA treatment.

Although rod-like MG express rod genes, they are, by no means, mature rods. First of all, they lack ROS and nuclear architecture of the mature rods. In addition, their gene expression profile was different from that of natural rod. We compared gene expression profiles of rod- like cells with natural rods and found that rod-like cells still lacked the expression of some genes related to neuron projection and photoreceptor cell cilium such as Syt1 and Cacnb2 (Supplementary Figure 10), indicating that rod-like cells were not fully differentiated. In addition, the rod-like MG still expressed MG genes such as Glul. By comparing differentially expressed genes associated with cell differentiation, we found that Notch-related genes, including Notch1 and its downstream gene Hes1, remain highly expressed in the reactivated and rod-like MG (Supplementary Figure 14B). Notch signaling pathway is involved in regulating both cell proliferation and cell fate specification during retinal development (Jadhav, Mason and Cepko, 2006; Yaron et al., 2006). Genetic removal of Notch1, or Notch downstream effectors such as Hes1 and RBP-J leads to precocious cell cycle exit and increased photoreceptors production during early retinogenesis (Tomita et al., 1996; Jadhav, Mason and Cepko, 2006; Riesenberg et al., 2009; Zheng et al., 2009). In addition, Notch1 functions in postmitotic cell fates to repress the rod fate (Mizeracka, Demaso and Cepko, 2013). Therefore, we conjecture that the high activity of Notch signaling may be the reason why rod-like cells fail to become full differentiated rods. A recent preprint has shown that suppressing Notch signaling in the MG and deletion of nuclear factor I factors a, b, and x (Nfia/b/x) synergize to promote neurogenesis of MG into retinal bipolar and amacrine interneurons in adult mice after NMDA-induced retinal injury (Le et al., 2023).

Limitations of the study

One unresolved question is how to further enhance rod regeneration by combining CCA with other factors. It remains to be tested whether inhibiting Notch signaling or overexpressing other neurogenic factors, in combination with CCA treatment, can result in the full differentiation of new rod photoreceptors. Additionally, while the CCA approach exhibited potent mitogenic activity in uninjured mouse retinas, its ability to promote MG proliferation after retinal injury is still unknown. Further investigation is needed to test the effects of CCA in mouse models of retinal degeneration. Finally, it is yet to be investigated how the CCA treatment affects the promoter accessibility of rod genes in the rod-like MG cells. Single-cell ATAC-seq or ChIP- seq analyses could provide insights into the epigenetic changes that underlie the rod-like phenotype acquired by the proliferated MG cells.

Materials and Methods

Animals

Wild-type mice (strain C57BL/6J), Rosa26-tdTomato reporter mice (strain B6.Cg- Gt(ROSA)26Sortm14(CAG-tdTomato)Hze/J), Rosa26-Sun1:GFP reporter mice (Strain B6.129- Gt(ROSA)26Sortm5.1(CAG-Sun1/sfGFP)Nat/MmbeJ), and Glast-CreER reporter mice (strain Tg(Slc1a3-cre/ERT)1Nat/J) mice were purchased from the Jackson Laboratory and were kept on a 12 hour light/12 hour dark cycle. All animal procedures performed were approved by the Hong Kong Department of Health under Animals Ordinance Chapter 340, and animal care was carried out in accordance with institutional guidelines.

AAV plasmids and vectors

pAAV-GFAP-GFP vector plasmid was cloned by replacing the CMV promoter with a 681bp ABC1D region of the human GFAP gene promoter (Lee et al., 2008) into the pAAV-CMV- GFP vector (Supplementary Figure 1A). cDNAs encoding mouse cyclin D1 were cloned into AAV plasmids by Gibson ligation. AAV-shRNA vectors were cloned by replacing the GFP sequence with mCherry-shRNA in the pAAV-GFAP-GFP vector. pAAV-GFAP-cyclinD1-P27 shRNA and pAAV-GFAP-GFP-LacZ shRNA were cloned by replacing mCherry sequence with cyclinD1 and GFP from the pAAV-GFAP-mCherry-P27 shRNA1 and pAAV-GFAP- mCherry-LacZ shRNA, respectively (Supplementary Figure 2A). pAAV rep/Cap 2/2, and Adenovirus helper plasmids were obtained from the University of Pennsylvania Vector Core, Philadelphia. pAAV7m8 plasmid (#64839) was purchased from addgene.

AAV was produced in HEK293T cells (HCL4517; Thermo Scientific) by AAV vector, rep/cap packaging plasmid, and adenoviral helper plasmid co-transfection followed by iodixanol gradient ultracentrifugation. Purified AAVs were concentrated with Amicon 100K columns (EMD Millipore) to a final titer higher than 5 x 1013 genome copies/mL. Protein gels were run to determine virus titers.

Tamoxifen injection

Intraperitoneal injections of tamoxifen (Sigma) in corn oil were administered to induce the expression of Cre recombinase. Starting from P20, tamoxifen was given at a dosage of 50 mg/kg daily for 5 consecutive days to label as many MG as possible. For the purpose of sparse labeling, a single dose of tamoxifen at 15 mg/kg was administered on P20.

AAV injection

Intravitreal injection was performed using a pulled angled glass pipette controlled by a FemtoJet (Eppendorf). The tip of the needle was passed through the sclera at the equator, near the dorsal limbus of the eyeball, and entered into the vitreous cavity. The injection volume of AAVs (5 x 1013 genome copies/mL) was 0.5μl per eye for P6 injection and 1μl per eye for adult mouse injection.

In situ RNA hybridization

In situ RNA hybridization was performed using the RNAscope Multiplex Fluorescent Detection Reagents V2 kit (Advanced Cell Diagnostics) following the commercial protocols. In brief, retinas were dissected, 4% PFA fixed, dehydrated in sucrose solution and embedded in optimal cutting temperature (OCT) medium. Retinas were cryosectioned into 20μm and mounted on SuperFrost Plus glass slides (Epredia). After OCT removal by PBS and further dehydration by ethanol, retinal sections were stained with GFP antibody (AB_2307313; Aves Labs) at 4 °C overnight. After three times wash with PBST (PBS with 0.1% Tween-20), retinal sections were hybridized with RNA probes (Advanced Cell Diagnostics Cat.No. 474801-C3 for Mm Rho, 524881-C2 for Mm Gnat1, 426231-C2 for Mm Glul) for 2 hours at 40 °C for two hours. Following in situ RNA hybridization steps, slides were stained using secondary antibodies (Jackson ImmunoResearch) and DAPI for two hours at room temperature. The fluorescent signals were visualized and captured using Nikon A1HD25 High speed and Large Field of View Confocal Microscope. The mRNA levels inside the GFP labelled nuclei membrane were quantified as number of RNA dots or measured as signal intensity level by ImageJ.

Immunohistochemistry

Retinas were dissected and fixed in 4% formaldehyde in PBS for 30 min at room temperature and sectioned at 20-μm thickness by cryostat. Retinal sections were blocked in 5% BSA in PBST (PBS with 0.1% Triton X-100), stained with primary antibodies at 4 °C overnight, and washed three times with PBST. Primary antibodies used in this study included rabbit anti- p27kip1(1:200, PA5-16717; Thermofisher); rabbit anti-cyclin D1 antibody (1:300, 26939-1-AP; Proteintech), goat anti-Otx2 antibody (1:200, AF 1979; R&D systems), mouse anti-HuC/D antibody (1:200, A21271; Thermofisher), rabbit anti-GFAP antibody (Z0334, 1:500, DAKO), rabbit anti-Sox2 antibody (1:2000, ab97959; abcam) and rabbit anti-Sox9 antibody (1:1000, AB5535; Millipore). Sections were stained using secondary antibodies (Jackson ImmunoResearch) for two hours at room temperature. Cell nuclei were counterstained with DAPI (Sigma). TUNEL was performed using an TUNEL BrightRed Apoptosis Detection Kit (Vazyme) according to the manufacturer’s protocol.

EdU incorporation and BrdU detection assay

5’-ethynyl-2’-deoxyuridine (EdU, 50mg/kg, Abcam ab146186) or 5-bromo-2’ deoxyuridine (BrdU,100mg/kg, Abcam 142567) was intraperitoneally injected to label the cells in the S phase. EdU staining was performed using the Click-iT™ EdU Alexa Fluor™ 488 Imaging Kit (C10337). For BrdU detection, the retinal sections were incubated with 2 M HCl for 1 hour at room temperature. The sections were rinsed with PBST and incubated with a blocking buffer containing 5% BSA in PBST for 2 hours at room temperature. Primary antibody mouse anti- BrdU (1:300, ab8152; Abcam) was stained overnight at 4 ℃, and secondary antibodies (Jackson ImmunoResearch) were stained for 2 hours at room temperature.

Single-cell RNA-seq

Library preparation and sequencing of single cells

The FACS-sorted cells from each sample were filtered through 40 µm strainer (pluriStrainer) and processed with Chromium Next GEM Single Cell 3′ GEM, Library & Gel Bead Kit v3.1 (1000269), Chromium Next GEM Chip G Single Cell Kit (1000127) and Chromium Controller (10x Genomics) according to the manufacturer’s protocol. The constructed libraries were sent to Novogene (Beijing) for NovaSeq paired-end 150 bp sequencing and produced 330 Gb of raw data.

Preprocessing, filtering and clustering of scRNA data

Sequencing results were preprocessed with 10x Genomics Cell Ranger 6.1.1. (Zheng et al., 2017) for demultiplexing, barcode assignment, and unique molecular identifier (UMI) quantification using a standard pipeline for 10X for each of the three samples. We used indexed mouse genome mm10 with added tdTomato and GFP transgenes as a reference. Reads mapping to introns covered more than 34% of all reads. Therefore, they were included in feature counts by using ‘include_introns’ option in cellranger count step of the pipeline. Cell Ranger aggregation was applied for merging runs from three samples.

Quality control, filtering, dimensional reduction, and clustering of the data were carried out using the Seurat package in R (Stuart et al., 2019). The cells with less than 2,000 expressed features, total UMI count higher than 100,000 and percentage of mitochondrial transcripts more than 15% were disregarded. Features of the Y chromosome, as well as Xist and Tsix genes of the X chromosome, were excluded to avoid confusing effects of gender.

Further filtering of cells was carried out by only including cells expressing tdTomato (Supplementary Figure 6). Potential cell duplicates were estimated using DoubletFinder method (McGinnis et al., 2019) implemented in R (with parameter settings PC=10, pK=0.28, pN=0.3). The threshold for the expected doublet formation rate was set to 10% to reflect the assumption that cell duplicates can appear relatively commonly in the MG cells. Detected duplicates were excluded from the data (Supplementary Figure 6G,H).

For the remaining cells, Uniform Manifold Approximation and Projection (UMAP) dimension reduction based on eight principal components (PCs) was applied, and cells were clustered using the graphical clustering method in Seurat. Cell types were identified using known marker genes (Clark et al., 2019; Hoang et al., 2020). After removing doublets, cell types in clusters 4-9 other than MG, Rod-like, and Rod included only a few cells, and 10-30% of the cells were from the control sample, suggesting that these cells are likely to be contaminated (Supplementary Figure 6I,J). After excluding cells in clusters 4-9, the new UMAP and clustering (Supplementary Figure 7A) revealed two subclusters of Rod-like cells (Clusters C3- C4, Supplementary Figure 7A). The cells in the subcluster C4 exhibited similar proportional contributions from all three groups and expressed lower level of rod-specific genes such as Rho (Supplementary Figure 7B,C). These cells were again defined as contaminated and excluded from the data. The final UMAP was constructed using seven PCs, and the same PCs were used for clustering of cells (Figure 3B).

Analysis of scRNA-seq data

The cell cycle state of each cell was defined by the Cell Cycle Scoring function in the Seurat R package. The cells with G2/M and S scores lower than 0.1 were defined to be proliferating (Supplementary Figure 7D,E). Differential gene expression analysis between different cell types or three treatment groups was performed using Deseq2 (Supplementary table 1) (Love et al., 2014). Enrichment analysis of top 200 up- and downregulated genes was performed for Gene Ontologies (GO), KEGG and Reactome pathways using Gprofiler2 R interface to g:Profiler (Raudvere et al., 2019).

The optomotor and electroretinography tests

Mouse visual acuity was measured using an Optometry System (Cerebral Mechanics Inc.) following the published protocol (Douglas et al., 2005). Testing was done with a grating of 12 degrees/second drifting speed and 100% contrast. The injected right eyes (CCA treated) and uninjected left eyes (Ctrl) were tested independently for counterclockwise and clockwise grating rotations, respectively. A staircase procedure was used, in which the observer tested low to high visual acuity. Each animal was tested for about 10-15 minutes per session.

ERG was measured using an Espion E3 System (Diagonsys LLC Inc.) as previously described (Hoang et al., 2023). Animals were dark-adapted overnight before the test. After inducing anesthesia with a ketamine: xylazaine injection intraperitoneally, the pupils of both eyes were dilated using Mydrin®-P Ophthalmic Solution. Once the pupils were fully dilated, the eyes were maintained moist using a topical gel. The measurement electrodes were then applied to the cornea of both eyes, while the ground electrodes were attached to the mouth and tail. All these steps were performed in the darkroom under dim red light. For scotopic ERG recordings, a multiple 530nm light with different intensities (increments from 0.01 cd.s/m2 to 30 cd.s/m2) were elicited to stimulate scotopic responses in a specific time interval. For photopic ERG recordings, 5 mins exposure under 10 cd.s/m2 light intensity was adopted to inhibit the rod function. The photopic response was measured by multiple flashes of 30 cd.s/m2 intensity in the illuminated background (10 cd.s/m2). The average amplitude and implicit time of a- and b- wave were recorded and exported for further analysis. The b-wave amplitude of the step-wise scotopic responses and that of photopic response at 30 cd.s/m2 were shown in figure.

Statistics

Data were represented as mean ± SEM in all figures. Sample sizes and statistical analysis were indicated for each experiment in figure legend. ANOVA analysis with Tukey test was performed to compare multiple groups, and Student’s t-test to compare two groups. A P value of P < 0.05 was considered statistically significant. GraphPad Prism was used to perform statistical analysis and make figures.

Supporting information

eLife Assessment

The manuscript presents a potentially important strategy to stimulate mammalian Müller glia to proliferate in vivo by manipulating cell cycle components. The findings are likely to appeal to retinal specialists and neuroscientists in general. However, the evidence that these cells become neurogenic is lacking/incomplete, suggesting that additional barriers exist to stimulate the regeneration of retinal neurons.

Reviewer #1 (Public Review):

Summary:

In this manuscript, Wu et al. introduce a novel approach to reactivate the Muller glia cell cycle in the mouse retina by simultaneously reducing p27Kip1 and increasing cyclin D1 using a single AAV vector. The approach effectively promotes Muller glia proliferation and reprograming without disrupting retinal structure or function. Interestingly, reactivation of the Muller glia cell cycle downregulates IFN pathway, which may contribute to the induced retinal regeneration. The results presented in this manuscript may offer a promising approach for developing Müller glia cell-mediated regenerative therapies for retinal diseases.

Strengths:

The data are convincing and supported by appropriate, validated methodology. These results are both technically and scientifically exciting and are likely to appeal to retinal specialists and neuroscientists in general.

Weaknesses:

There are some data gaps that need to be addressed.

(1) Please label the time points of AAV injection, EdU labeling, and harvest in Figure 1B.

(2) What fraction of Müller cells were transduced by AAV under the experimental conditions?

(3) It seems unusually rapid for MG proliferation to begin as early as the third day after CCA injection. Can the authors provide evidence for cyclin D1 overexpression and p27 Kip1 knockdown three days after CCA injection?

(4) The authors reported that MG proliferation largely ceased two weeks after CCA treatment. While this is an interesting finding, the explanation that it might be due to the dilution of AAV episomal genome copies in the dividing cells seems far-fetched.

Reviewer #2 (Public Review):

This manuscript by Wu, Liao et al. reports that simultaneous knockdown of P27Kip1 with overexpression of Cyclin D can stimulate Muller glia to re-enter the cell cycle in the mouse retina. There is intense interest in reprogramming mammalian muller glia into a source for neurogenic progenitors, in the hopes that these cells could be a source for neuronal replacement in neurodegenerative diseases. Previous work in the field has shown ways in which mouse Muller glia can be neurogenically reprogrammed and these studies have shown cell cycle re-entry prior to neurogenesis. In other works, typically, the extent of glial proliferation is limited, and the authors of this study highlight the importance of stimulating large numbers of Muller glia to re-enter the cell cycle with the hopes they will differentiate into neurons. While the evidence for stimulating proliferation in this study is convincing, the evidence for neurogenesis in this study is not convincing or robust, suggesting that stimulating cell cycle-reentry may not be associated with increasing regeneration without another proneural stimulus.

Below are concerns and suggestions.

Intro:

(1) The authors cite past studies showing "direct conversion" of MG into neurons. However, these studies (PMID: 34686336; 36417510) show EdU+ MG-derived neurons suggesting cell cycle re-entry does occur in these strategies of proneural TF overexpression.

(2) Multiple citations are incorrectly listed, using the authors first name only (i.e. Yumi, et al; Levi, et al;). Studies are also incompletely referenced in the references.

Figure 1:

(3) When are these experiments ending? On Figure 1B it says "analysis" on the end of the paradigm without an actual day associated with this. This is the case for many later figures too. The authors should update the paradigms to accurately reflect experimental end points.

(4) Are there better representative pictures between P27kd and CyclinD OE, the EdU+ counts say there is a 3 fold increase between Figure 1D&E, however the pictures do not reflect this. In fact, most of the Edu+ cells in Figure 1E don't seem to be Sox9+ MG but rather horizontally oriented nuclei in the OPL that are likely microglia.

(5) Is the infection efficacy of these viruses different between different combinations (i.e. CyclinD OE vs. P27kd vs. control vs. CCA combo)? As the counts are shown in Figure 1G only Sox9+/Edu+ cells are shown not divided by virus efficacy. If these are absolute counts blind to where the virus is and how many cells the virus hits, if the virus efficacy varies in efficiency this could drive absolute differences that aren't actually biological.

(6) According to the Jax laboratories, mice aren't considered aged until they are over 18months old. While it is interesting that CCA treatment does not seem to lose efficacy over maturation I would rephrase the findings as the experiment does not test this virus in aged retinas.

(7) Supplemental Figure 2c-d. These viruses do not hit 100% of MG, however 100% of the P27Kip staining is gone in the P27sh1 treatment, even the P27+ cell in the GCL that is likely an astrocyte has no staining in the shRNA 1 picture. Why is this?

Figure 2

(8) Would you expect cells to go through two rounds of cell cycle in such a short time? The treatment of giving Edu then BrdU 24 hours later would have to catch a cell going through two rounds of division in a very short amount of time. Again the end point should be added graphically to this figure.

Figure 3

(9) I am confused by the mixing of ratios of viruses to indicate infection success. I know mixtures of viruses containing CCA or control GFP or a control LacZ was injected. Was the idea to probe for GFP or LacZ in the single cell data to see which cells were infected but not treated? This is not shown anywhere?

(10) The majority of glia sorted from TdTomato are probably not infected with virus. Can you subset cells that were infected only for analysis? Otherwise it makes it very hard to make population judgements like Figure 3E-H if a large portion are basically WT glia.

(11) Figure 3C you can see Rho is expressed everywhere which is common in studies like this because the ambient RNA is so high. This makes it very hard to talk about "Rod-like" MG as this is probably an artifact from the technique. Most all scRNA-seq studies from MG-reprogramming have shown clusters of "rods" with MG hybrid gene expression and these had in the past just been considered an artifact.

(12) It is mentioned the "glial" signature is downregulated in response to CCA treatment. Where is this shown convincingly? Figure H has a feature plot of Glul , which is not clear it is changed between treatments. Otherwise MG genes are shown as a function of cluster not treatment.

Figure 4

(13) The authors should be commended for being very careful in their interpretations. They employ the proper controls (Er-Cre lineage tracing/EdU-pulse chasing/scRNA-seq omics) and were very careful to attempt to see MG-derived rods. This makes the conclusion from the FISH perplexing. The few puncta dots of Rho and GNAT in MG are not convincing to this reviewer, Rho and GNAT dots are dense everywhere throughout the ONL and if you drew any random circle in the ONL it would be full of dots. The rigor of these counts also comes into question because some dots are picked up in MG in the INL even in the control case. This is confusing because baseline healthy MG do not express RNA-transcripts of these Rod genes so what is this picking up? Taken together, the conclusion that there are Rod-like MG are based off scRNA-seq data (which is likely ambient contamination) and these FISH images. I don't think this data warrants the conclusion that MG upregulate Rod genes in response to CCA.

Figure 5

(14) Similar point to above but this Glul probe seems odd, why is it throughout the ONL but completely dark through the IPL, this should also be in astrocytes can you see it in the GCL? These retinas look cropped at the INL where below is completely black. The whole retinal section should be shown. Antibodies exist to GS that work in mouse along with many other MG genes, IHC or western blots could be done to better serve this point.

Figure 6

(15) Figure 6D is not a co-labeled OTX2+/ TdTomato+ cell, Otx2 will fill out the whole nucleus as can be seen with examples from other MG-reprogramming papers in the field (Hoang, et al. 2020; Todd, et al. 2020; Palazzo, et al. 2022). You can clearly see in the example in Figure 6D the nucleus extending way beyond Otx2 expression as it is probably overlapping in space. Other examples should be shown, however, considering less than 1% of cells were putatively Otx2+, the safer interpretation is that these cells are not differentiating into neurons. At least 99.5% are not.

(16) Same as above Figure 6I is not convincingly co-labeled HuC/D is an RNA-binding protein and unfortunately is not always the clearest stain but this looks like background haze in the INL overlapping. Other amacrine markers could be tested, but again due to the very low numbers, I think no neurogenesis is occurring.

(17) In the text the authors are accidently referring to Figure 6 as Figure 7.

Figure 7

(18) I like this figure and the concept that you can have additional MG proliferating without destroying the retina or compromising vision. This is reminiscent of the chick MG reprogramming studies in which MG proliferate in large numbers and often do not differentiate into neurons yet still persist de-laminated for long time points.

General:

(19) The title should be changed, as I don't believe there is any convincing evidence of regeneration of neurons. Understanding the barriers to MG cell-cycle re-entry are important and I believe the authors did a good job in that respect, however it is an oversell to report regeneration of neurons from this data.

(20) This paper uses multiple mouse lines and it is often confusing when the text and figures switch between models. I think it would be helpful to readers if the mouse strain was added to graphical paradigms in each figure when a different mouse line is employed.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In this manuscript, Wu et al. introduce a novel approach to reactivate the Muller glia cell cycle in the mouse retina by simultaneously reducing p27Kip1 and increasing cyclin D1 using a single AAV vector. The approach effectively promotes Muller glia proliferation and reprograming without disrupting retinal structure or function. Interestingly, reactivation of the Muller glia cell cycle downregulates IFN pathway, which may contribute to the induced retinal regeneration. The results presented in this manuscript may offer a promising approach for developing Müller glia cell-mediated regenerative therapies for retinal diseases.

Strengths:

The data are convincing and supported by appropriate, validated methodology. These results are both technically and scientifically exciting and are likely to appeal to retinal specialists and neuroscientists in general.

Weaknesses:

There are some data gaps that need to be addressed.

(1) Please label the time points of AAV injection, EdU labeling, and harvest in Figure 1B.

We thank the reviewer for highlighting the lack of clarity in our experimental design. We have labeled all experiment timelines in the figures where appropriate in the revised version.

(2) What fraction of Müller cells were transduced by AAV under the experimental conditions?

We apologize for not clearly explaining the AAV transduction effeciency. AAV transduction efficiency was not uniform across the retinas. The retinal region adjacent to the optic nerve exhibits a transduction efficiency of nearly 100%. In contrast, the peripheral retina shows a lower transduction efficiency compared to the central region. The representative retinal sections with typical infection pattern are shown in Supplementary figure 4. The quantification of Edu+ MG or other markers was conducted in a 250 µm region with the highest efficiency. For scRNA-seq experiment, retinal regions with high AAV transduction efficiency were dissected with the aid of a control GFP virus.

(3) It seems unusually rapid for MG proliferation to begin as early as the third day after CCA injection. Can the authors provide evidence for cyclin D1 overexpression and p27 Kip1 knockdown three days after CCA injection?

We included the data that GFP expression is evident at 3 days post AAV-GFP-GFP injection (Supplementary Fig. 1B). Additionally, we performed immunostaining and confirmed cyclin D1 overexpression at 3 days post CCA injection (Fig. 2E) as well as qPCR analysis to confirm cyclin D1 overexpression and p27kip1 knockdown at the same time point (Supplementary Fig. 5).

(4) The authors reported that MG proliferation largely ceased two weeks after CCA treatment. While this is an interesting finding, the explanation that it might be due to the dilution of AAV episomal genome copies in the dividing cells seems far-fetched.

We agree with the reviewer that dilution of AAV episomal genomes is unlikely to be the sole reason for the stop of MG proliferation. By staining cyclin D1 at various days post CCA injection, we found that cyclin D1 is immediately downregulated in the mitotic MG undergoing interkinetic nuclear migration to the outer nuclear layer (Fig. 2G-I). In contrast, the effect of p27kip1 knockdown by CCA lasted longer (Supplementary Figure 9-10). It is possible that other anti-proliferative genes are involved in the immediate downregulation of Cyclin D1.

Reviewer #2 (Public Review):

This manuscript by Wu, Liao et al. reports that simultaneous knockdown of P27Kip1 with overexpression of Cyclin D can stimulate Muller glia to re-enter the cell cycle in the mouse retina. There is intense interest in reprogramming mammalian muller glia into a source for neurogenic progenitors, in the hopes that these cells could be a source for neuronal replacement in neurodegenerative diseases. Previous work in the field has shown ways in which mouse Muller glia can be neurogenically reprogrammed and these studies have shown cell cycle re-entry prior to neurogenesis. In other works, typically, the extent of glial proliferation is limited, and the authors of this study highlight the importance of stimulating large numbers of Muller glia to re-enter the cell cycle with the hopes they will differentiate into neurons. While the evidence for stimulating proliferation in this study is convincing, the evidence for neurogenesis in this study is not convincing or robust, suggesting that stimulating cell cycle-reentry may not be associated with increasing regeneration without another proneural stimulus.

Below are concerns and suggestions.

Intro:

(1) The authors cite past studies showing "direct conversion" of MG into neurons. However, these studies (PMID: 34686336; 36417510) show EdU+ MG-derived neurons suggesting cell cycle re-entry does occur in these strategies of proneural TF overexpression.

We thank the reviewer for pointing this out. We have revised the statement to "MG reprogramming".

(2) Multiple citations are incorrectly listed, using the authors first name only (i.e. Yumi, et al; Levi, et al;). Studies are also incompletely referenced in the references.

We apologize for the mistakes in reference. We have corrected the reference mistakes in the revised version.

Figure 1:

(3) When are these experiments ending? On Figure 1B it says "analysis" on the end of the paradigm without an actual day associated with this. This is the case for many later figures too. The authors should update the paradigms to accurately reflect experimental end points.

We thank the reviewer for highlighting the lack of clarity in our experimental design. We have labeled all experiment timelines in the figures where appropriate in the revised version.

(4) Are there better representative pictures between P27kd and CyclinD OE, the EdU+ counts say there is a 3 fold increase between Figure 1D&E, however the pictures do not reflect this. In fact, most of the Edu+ cells in Figure 1E don't seem to be Sox9+ MG but rather horizontally oriented nuclei in the OPL that are likely microglia.

Thanks to the reviewer for pointing this out. We have replaced the image of cyclin D1 OE retina which a more representative image.

(5) Is the infection efficacy of these viruses different between different combinations (i.e. CyclinD OE vs. P27kd vs. control vs. CCA combo)? As the counts are shown in Figure 1G only Sox9+/Edu+ cells are shown not divided by virus efficacy. If these are absolute counts blind to where the virus is and how many cells the virus hits, if the virus efficacy varies in efficiency this could drive absolute differences that aren't actually biological.

Rule out the possibility that the differences in MG proliferation across groups are due to variations in viral efficacy, we have examined the p27kip1 knockdown and cyclin D1 overexpression efficiencies for all four groups by qPCR analysis. The result showed that cyclin D1 overexpression efficiency by AAV-GFAP-Cyclin D1 virus alone or P27 knockdown efficiency by AAV-GFAP-mCherry-p27kip1 shRNA1 is comparable to, if not even higher than, those by CCA virus (Supplementary Fig 5). Therefore, the virus efficacy cannot explain the drastic increase in MG proliferation by CCA.

As the central retina usually had 100% infection efficacy (Supplementary Fig. 4), we quantified the Edu+Sox9+ cell number in the 250µm regions next to the optic nerve.

(6) According to the Jax laboratories, mice aren't considered aged until they are over 18months old. While it is interesting that CCA treatment does not seem to lose efficacy over maturation I would rephrase the findings as the experiment does not test this virus in aged retinas.

Thank you to the reviewer for bringing this to our attention. We have changed to “older adult mice” in our revised manuscript.

(7) Supplemental Figure 2c-d. These viruses do not hit 100% of MG, however 100% of the P27Kip staining is gone in the P27sh1 treatment, even the P27+ cell in the GCL that is likely an astrocyte has no staining in the shRNA 1 picture. Why is this?

We have replaced the images in Supplementary Fig. 2B-D.

Figure 2

(8) Would you expect cells to go through two rounds of cell cycle in such a short time? The treatment of giving Edu then BrdU 24 hours later would have to catch a cell going through two rounds of division in a very short amount of time. Again the end point should be added graphically to this figure.

We thank the reviewer for the comment. We repeated the Edu/BrdU colabelling experiment with extended periods of Edu/BrdU injections. Based on the result of the MG proliferation time course study (Fig. 2A), we injected 5 times of Edu from D1 to D5 and 5 times of BrdU from D6 to D10 post-CCA injection, which covered the major phase of MG proliferation (Fig. 2B-C). Consistent with the previous findings, we did not observe any BrdU&EdU double positive MG cells.

Additionally, we showed that cyclin D1 overexpression immediately ceased in migrating mitotic MG (Fig. 2G-I), which may explain why CCA-treated MG do not progress to the second round of cell division.

Figure 3

(9) I am confused by the mixing of ratios of viruses to indicate infection success. I know mixtures of viruses containing CCA or control GFP or a control LacZ was injected. Was the idea to probe for GFP or LacZ in the single cell data to see which cells were infected but not treated? This is not shown anywhere?

The virus infection was not uniform across the entire retina (Supplementary Fig. 4). To mark the infection hotspots, we added 10% GFP virus to the mixture. Regions of the retina with low infection efficiency were removed by dissection and excluded from the scRNA-seq analysis. Therefore, we assumed that the vast majority of MG were infected by CCA. We apologize for not clearly explaining this methodological detail in the original text. We have added the experimental design to Fig. 3A and revised the result part (line 191-196) accordingly.

(10) The majority of glia sorted from TdTomato are probably not infected with virus. Can you subset cells that were infected only for analysis? Otherwise it makes it very hard to make population judgements like Figure 3E-H if a large portion are basically WT glia.

This question is related to the last one. Since the regions with high virus infection efficiency were selectively dissected and isolated for analysis, the CCA-infected MG should constitute the vast majority of MG in the scRNA-seq data.

(11) Figure 3C you can see Rho is expressed everywhere which is common in studies like this because the ambient RNA is so high. This makes it very hard to talk about "Rod-like" MG as this is probably an artifact from the technique. Most all scRNA-seq studies from MG-reprogramming have shown clusters of "rods" with MG hybrid gene expression and these had in the past just been considered an artifact.

We agree with the reviewer that the high rod gene expression in the rod-MG cluster is an artifact. We have performed multiple rounds of RNA in situ hybridization on isolated MG nuclei. The counts of Gnat1 and Rho mRNA signal are largely overlapped between the two samples with and without CCA treatment (Supplementary Fig 14). Some MG in the control retinas without CCA treatment had up to 7 or 8 dots per cell, suggesting contamination of attached rod cell debris during retina dissociation (Supplementary Fig 14). Therefore, the result did not support that rod-MG is a reprogrammed MG population with rod gene upregulation.

(12) It is mentioned the "glial" signature is downregulated in response to CCA treatment. Where is this shown convincingly? Figure H has a feature plot of Glul, which is not clear it is changed between treatments. Otherwise MG genes are shown as a function of cluster not treatment.

We have added box plots of several MG-specific genes to illustrate the downregulation of the glial signature in the relevant cell cluster in the revised manuscript (Supplementary Fig. 15).

Figure 4

(13) The authors should be commended for being very careful in their interpretations. They employ the proper controls (Er-Cre lineage tracing/EdU-pulse chasing/scRNA-seq omics) and were very careful to attempt to see MG-derived rods. This makes the conclusion from the FISH perplexing. The few puncta dots of Rho and GNAT in MG are not convincing to this reviewer, Rho and GNAT dots are dense everywhere throughout the ONL and if you drew any random circle in the ONL it would be full of dots. The rigor of these counts also comes into question because some dots are picked up in MG in the INL even in the control case. This is confusing because baseline healthy MG do not express RNA-transcripts of these Rod genes so what is this picking up? Taken together, the conclusion that there are Rod-like MG are based off scRNA-seq data (which is likely ambient contamination) and these FISH images. I don't think this data warrants the conclusion that MG upregulate Rod genes in response to CCA.

Given the results of RNA in situ hybridization on isolated MG, we revisited the result of the RNA in situ hybridization on retinal sections as well. We performed RNA in situ in the retinal section at 1 week post CCA treatment, expecting to see lower Gnat1 and Rho signals in the ONL-localizing MG compared to 3 weeks and 4 months post CCA treatment. However, we observed similar levels across all three time points (data not shown). The lack of dynamic changes in rod gene expression levels also suggests contamination from tightly surrounding neighboring rods. Consequently, we have reinterpreted the scRNA-seq and RNA FISH data and withdrawn the conclusion that MG upregulated rod genes after CCA treatment. We thank the reviewer for pointing out this potential issue and helping us avoid an incorrect conclusion.

Figure 5

(14) Similar point to above but this Glul probe seems odd, why is it throughout the ONL but completely dark through the IPL, this should also be in astrocytes can you see it in the GCL? These retinas look cropped at the INL where below is completely black. The whole retinal section should be shown. Antibodies exist to GS that work in mouse along with many other MG genes, IHC or western blots could be done to better serve this point.

We have replaced the images in Figure 4 in the revised manuscript. Additionally, we have performed the Sox9 antibody staining to demonstrate partial MG dedifferentiation following CCA treatment (Figure 5).

Figure 6

(15) Figure 6D is not a co-labeled OTX2+/ TdTomato+ cell, Otx2 will fill out the whole nucleus as can be seen with examples from other MG-reprogramming papers in the field (Hoang, et al. 2020; Todd, et al. 2020; Palazzo, et al. 2022). You can clearly see in the example in Figure 6D the nucleus extending way beyond Otx2 expression as it is probably overlapping in space. Other examples should be shown, however, considering less than 1% of cells were putatively Otx2+, the safer interpretation is that these cells are not differentiating into neurons. At least 99.5% are not.

We have replaced the image of Otx2+ Tdt+ Edu+ cell, which shows the whole nucleus filled with strong Otx2 staining.

(16) Same as above Figure 6I is not convincingly co-labeled HuC/D is an RNA-binding protein and unfortunately is not always the clearest stain but this looks like background haze in the INL overlapping. Other amacrine markers could be tested, but again due to the very low numbers, I think no neurogenesis is occurring.

Since we didn’t find HuC/D+Tdt+EdU+ cells at 3 weeks post CCA treatment, we believe that the weak HuC/D+ staining in the MG daughter cells at 4 months is not background, but rather reflects an incomplete neurogenic switch. This suggests that the process of neurogenesis may be ongoing but not fully realized within the observed timeframe without additional stimuli.

(17) In the text the authors are accidently referring to Figure 6 as Figure 7.

We thank the reviewer for pointing out the mistake. We will correct the mistake in the revised manuscript.

Figure 7

(18) I like this figure and the concept that you can have additional MG proliferating without destroying the retina or compromising vision. This is reminiscent of the chick MG reprogramming studies in which MG proliferate in large numbers and often do not differentiate into neurons yet still persist de-laminated for long time points.

General:

(19) The title should be changed, as I don't believe there is any convincing evidence of regeneration of neurons. Understanding the barriers to MG cell-cycle re-entry are important and I believe the authors did a good job in that respect, however it is an oversell to report regeneration of neurons from this data.

We thank the reviewer for the suggestion. We have changed the title to “Simultaneous cyclin D1 overexpression and p27kip1 knockdown enable robust Müller glia cell cycle reactivation in uninjured mouse retina” in the revised manuscript.

(20) This paper uses multiple mouse lines and it is often confusing when the text and figures switch between models. I think it would be helpful to readers if the mouse strain was added to graphical paradigms in each figure when a different mouse line is employed.

We have labeled the mouse lines used in each experiment in the figures where appropriate.

Version 2

Abstract

Abstract

Harnessing the regenerative potential of endogenous stem cells to restore lost neurons is a promising strategy for treating neurodegenerative disorders. Müller glia (MG), the primary glial cell type in the retina, exhibit extraordinary regenerative abilities in zebrafish, proliferating and differentiating into neurons post-injury. However, the regenerative potential of mouse MG is limited by their inherent inability to re-enter the cell cycle, constrained by high levels of the cell cycle inhibitor p27Kip1 and low levels of cyclin D1. Here, we report a method to drive robust MG proliferation by adeno-associated virus (AAV)-mediated cyclin D1 overexpression and p27Kip1 knockdown. MG proliferation induced by this dual targeting vector was self-limiting, as MG did not undergo uncontrolled proliferation. As shown by single-cell RNA-sequencing, cell cycle reactivation led to suppression of interferon signaling, activation of reactive gliosis, and downregulation of glial genes in MG. Over time, the majority of the MG daughter cells retained the glial fate, resulting in an expanded MG pool. Interestingly, about 1% MG daughter cells expressed markers for retinal interneurons, suggesting latent neurogenic potential in a small MG subset. By establishing a safe, controlled method to promote MG proliferation in vivo while preserving retinal integrity, this work provides a valuable tool for combinatorial therapies integrating neurogenic stimuli to promote neuron regeneration.

Introduction

Müller glia (MG) are the last cell type generated by the retinal progenitor cells (RPCs) during development and exhibit a gene expression profile similar to that of late retinal progenitor cells (Jadhav et al., 2009; Roesch et al., 2012). In teleost fish and amphibians, MG respond rapidly to retinal injury, undergoing robust proliferation and regenerating lost retinal neurons from MG-derived progenitor cells (Hamon et al., 2016; Todd and Reh, 2022). In contrast, the proliferative and neurogenic ability of MG in response to injury in mammals is severely limited, failing to mediate retinal self-repair (Karl et al., 2008). Recent investigations have achieved great success in mammalian retinal regeneration by stimulating MG reprogramming through a single or combination of neurogenic transcription factors (Jorstad et al., 2017; Todd et al., 2021, 2022; Ueki et al., 2015). However, this may lead to a depletion of the MG population and cause further retinal degeneration, as MG are indispensable for retinal function and homeostasis. Other studies have shown that the quiescent state of MG can be overridden by upstream signaling pathways such as Wnt and Hippo (Yao et al., 2016; Hamon et al., 2019; Rueda et al., 2019). Activation of Wnt signaling by forced expression of β-catenin in adult mouse MG promoted spontaneous cell cycle re-entry in uninjured retinas (Yao et al., 2016). Moreover, it was demonstrated that bypassing the Hippo pathway in mouse MG led to spontaneous re-entry into the cell cycle and reprogramming into a progenitor cell-like state (Hamon et al., 2019; Rueda et al., 2019). These findings suggest that the re-entry of MG into the cell cycle, which is the first step of MG-mediated retinal regeneration in zebrafish, could be unlocked in mammalian retinas.

The cell cycle of MG is mainly regulated by cyclins and cyclin-dependent kinases (CDKs). Cyclins bind to CDKs to form cyclin-CDK complexes to promote cell cycle progression. During retinal development, the expression of D-type cyclins (cyclin D1, D2, and D3) is tightly regulated (Barton and Levine, 2008; Dyer and Cepko, 2001; Trimarchi et al., 2008). Among these, cyclin D1, encoded by the Ccnd1 gene, is the predominant D-type cyclin in the developing retina and is highly expressed in the RPCs but absent in differentiated cells (Barton and Levine, 2008; Trimarchi et al., 2008). Mice lacking Ccnd1 have small eyes and hypocellular retinas due to reduced RPC proliferation (Fantl et al., 1995; Sicinski et al., 1995), which cannot be compensated by Ccnd2 and Ccnd3 (Das et al., 2012, 2009). Negative regulators of the cell cycle are CDK inhibitors (CDKIs), which include the INK4 (p16INK4a, p15INK4b, p18INK4c, and p19INK4d) and CIP/KIP families (p21Cip1, p27Kip1, and p57Kip2) (Reynisdóttir et al., 1995). The CDKIs inhibit cell cycle progression by binding to and inactivating the cyclin-CDK complexes (Besson et al., 2008), and they regulate the proliferation in distinct retinal progenitor cells (Dyer and Cepko, 2000, 2001; Levine et al., 2000). P27Kip1 inhibits the cyclin D-CDK complex to enter the S phase. Following acute retinal damage in mice, a very small number of MG re-enter the cell cycle, co-incident with the downregulation of p27Kip1 or upregulation of cyclin D1 (Dyer and Cepko, 2001; Hamon et al., 2019; Rueda et al., 2019; Yao et al., 2016). However, this process is transient, as cyclin D1 expression rapidly returns to the basal level (Hamon et al., 2019; Rueda et al., 2019).

In this study, we propose that targeting the two key regulators of the cell cycle, cyclin D1 and p27kip1, can effectively induce MG cell cycle re-entry. Our findings demonstrate that simultaneously reducing p27Kip1 and increasing cyclin D1 in MG using a single AAV vector has a strong synergistic effect on promoting MG proliferation in uninjured adult mouse retinas. MG proliferation induced by this treatment is robust and self-limiting, as MG undergo a single round of cell division rather than unlimited proliferations. Through single-cell RNA sequencing (scRNA-seq), we observed that cell cycle reactivation leads to the downregulation of the interferon pathways in the MG and suppression of the MG genes. By RNA in situ hybridization and immunostaining, we showed that MG partially and temporally suppressed their glial cell fate, while the majority of MG regained the normal glial identify by four months post-CCA treatment. A few EdU+ MG daughter cells in the inner nuclear layer (INL) express the bipolar cell marker Otx2 or the amacrine cell marker HuC/D, suggesting rare de novo neurogenesis from MG. Importantly, MG cell cycle reactivation does not disrupt retinal structure or impair retinal function, as the treatment did not deplete the MG from the retina or cause neoplasia. In summary, our results showed that MG cell cycle reactivation by downregulating p27Kip1 and upregulating cyclin D1 stimulates MG proliferation, and it is possible to combine this approach with other factors that promote regeneration to enhance retinal repair mediated by MG.

Results

Simultaneous p27Kip1 downregulation and cyclin D1 overexpression drive robust MG proliferation in the uninjured mouse retina

To test the hypothesis that adult mouse MG are kept in a quiescent state by high levels of p27Kip1 and low levels of cyclin D1, we examined whether spontaneous MG proliferation could be activated by directly changing the levels of these two downstream cell cycle regulators. To drive MG-specific gene expression, adeno-associated virus (AAV)-mediated transgene expression is controlled by a promoter sequence cloned from the human glial fibrillary acidic protein (GFAP) gene (Fig. 1A, Supplementary Fig. 1A) (Lee et al., 2008). AAV serotype 7m8 vectors were injected intravitreally into the eyes of C57BL/6 mice on postnatal day 6 (P6), when the majority of MG finish proliferation and start differentiation, and GFP reporter expression could be detected at three days post-AAV injection (Supplementary Fig. 1B). Our close examination showed that this GFAP promoter drives transgene expression specifically in MG but not in astrocytes (Supplementary Fig. 1C-F).

Simultaneous p27Kip1 downregulation and cyclin D1 overexpression drive robust MG proliferation in the uninjured mouse retina

(A) Schematic representations of AAV vectors used in this study. AAV-GFAP-GFP-non-target (NT) shRNA for control, AAV-GFAP-mCherry-p27 shRNA for p27kip1 knockdown (KD), AAV-GFAP-cyclin D1 for cyclin D1 overexpression (OE), and AAV-GFAP-cyclin D1-p27 shRNA for p27kip1 KD and cyclin D1 OE. (B) Experimental design. Mice received an intravitreal AAV injection on postnatal day 6 (P6), designated as D0, and daily EdU injections intraperitoneally from D7 to D11. (C-F) Analysis of EdU incorporation with Sox9 colabeling in uninjured mouse eyes injected with indicated viruses. ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer. (G) Quantification of EdU+Sox9+ cells per 250 µm high infection areas. Control (n=8 eyes), p27kip1 KD (n=11 eyes), cyclin D1 OE (n=8 eyes), CCA (n=14 eyes). (H) Quantification of EdU+Sox9+ cells per 250 µm high infection areas in retinas injected with CCA at indicated ages: P6 (n=13), P28 (n=17), and P255-P347 (n=7). Five EdU injections were given intraperitoneally from D7 to D11 post-CCA injection, and retinas were harvested for EdU analysis on D12. Data are presented as mean ± SEM. * P<0.05, ** P<0.01, *** P<0.001, ns = not significant (one-way ANOVA with Tukey’s post-hoc test) (G-H).

To examine the effect of p27Kip1 and cyclin D1 levels on MG proliferation, wild type mice were infected with control AAV or AAV that knocked down p27Kip1 and/or overexpressed cyclin D1 at P6 and received intraperitoneal injections of 5-ethynyl-2’-deoxyurdine (EdU) for five consecutive days, from P13 to P17, to label the MG that had entered the S phase of the cell cycle (Fig. 1A-B). In the control retinas that were infected by AAV7m8-GFAP-GFP-non-target (NT) shRNA, all MG, which were positive of Sox9 expression, were negative of EdU labeling (Fig. 1C). When the retina was infected by AAV7m8-GFAP-mCherry-p27Kip1 shRNA1, which expresses a highly efficient p27Kip1 shRNA1, a small number of MG cells re-entered the cell cycle (Fig. 1D, G, Supplementary Fig. 2); however, the vast majority of MG remained in a quiescent state. Overexpressing cyclin D1 alone through AAV7m8-GFAP-cyclinD1 infection stimulated a subset of MG cells to proliferate, resulting in a three-fold increase in the number of EdU+ MG cells compared to p27Kip1 knockdown (Fig. 1E, G, Supplementary Fig. 3). Due to the uneven efficiency of AAV infection across the retina, quantification was performed in the region with the highest infection levels (Supplementary Fig. 4). Finally, the AAV7m8-GFAP-cyclinD1-p27Kip1 shRNA1 vector, which simultaneously overexpressed cyclin D1 and suppressed p27Kip1, had the most significant impact on MG proliferation, with a five-fold increase in EdU+Sox9+ cells compared to cyclin D1 overexpression alone (Fig. 1F, G). The differences in MG proliferation across groups were not due to variations in virus infection efficacy, as confirmed by measuring p27Kip1 knockdown and cyclin D1 overexpression levels by quantitative PCR (Supplementary Fig. 5). A transgenic mouse line Glast-CreERT2; Sun1:GFP, in which MG nuclei were labeled by nuclear membrane-bound GFP (Supplementary Fig. 6A-C), was used to quantify the percentage of MG that re-entered the cell cycle. In the retinal area where viral infection rate was the highest, approximately 45% of Sun1:GFP+ MG were EdU positive, and the total number of MG increased by about 50% (Supplementary Fig. 6D-H), indicating that nearly half of the MG cells re-entered the cell cycle. As the AAV7m8-GFAP-cyclinD1-p27Kip1 shRNA1 vector enables such robust MG proliferation, we refer to it as the cell cycle activator (CCA) for short.

Previous research has shown that the ability of retina to regenerate by various stimuli declines with the age of the mice (Löffler et al., 2015; Ueki et al., 2015). We compared the efficiency of CCA in driving MG proliferation in young (P28) and older adult mouse (P255-P347) retinas to that in P6 pups (Fig. 1H). Remarkably, MG proliferation induced by CCA remained robust in adult mice (Fig. 1H). These findings suggest that CCA efficiently drives MG proliferation irrespective of the age of the mice. The synergistic effect of cyclin D1 overexpression and p27Kip1 knockdown on MG proliferation suggest that both low levels of p27Kip1 and high levels of cyclin D1 are required for postmitotic MG to re-enter the cell cycle.

MG proliferation driven by CCA is self-limiting

Concerned that p27Kip1 suppression and cyclin D1 overexpression may lead to uncontrolled cell proliferation and retinal tumorigenesis, we analysed the proliferative capacity of MG driven by CCA. Firstly, we examined the duration of MG proliferation after CCA treatment by a time-course EdU incorporation assay. Starting on various days after the CCA treatment, mice received two injections of EdU to label the cells that were undergoing proliferation (Fig. 2A). The result revealed that MG proliferation started as quickly as the third day after CCA injection, reached its peak around the fifth day, followed by a gradual decrease (Fig. 2A). By two weeks post-CCA injection, only a few MG were observed re-entering the cell cycle. Two months post-CCA injection, MG proliferation had mostly ceased (Fig. 2A). These findings suggest that MG proliferation was largely completed within two weeks post-CCA treatment. In addition, an EdU/BrdU double-labeling assay was performed to examine whether MG undergoes one or multiple cell divisions after CCA treatment. Five injections of EdU were given from day one to five post-CCA treatment, followed by five BrdU injections from day six to ten post-CCA treatment (Fig. 2B). Retinas were collected one day after the last BrdU injection to evaluate if any MG continuously entered the S phase of the cell cycle. While there were a number of cells positive for EdU or BrdU, no cells were co-labeled with EdU and BrdU (Fig. 2B,C), indicating that no MG underwent two cell divisions. Finally, we utilized the Glast-CreERT2; tdTomato mouse line to label MG sparsely with a low dose of tamoxifen induction (Fig. 2D, Supplementary Fig. 7). At four weeks post-CCA injection, we observed either single MG or pairs of MG in close proximity, but no clusters of three or more cells (Fig. 2E-F). The results of these experiments suggest that MG undergo only one cell division following CCA treatment.

MG proliferation induced by CCA is self-limiting

(A) Time-course analysis of MG proliferation following CCA injection. EdU was administered for two consecutive days, starting at various days post-CCA injection, with samples harvested one day after the second EdU injection. Data are presented as mean ± SEM (n≥4). (B) Analysis of cells labeled with EdU and BrdU. (C) Quantification of the percentages of EdU+BrdU−, EdU−BrdU+ and EdU+BrdU+ cells of the total Sox9+ cells. Data are presented as mean ± SEM (n=3). ***P<0.001, ****P<0.0001 (one-way ANOVA with Tukey post hoc test). (D) Experimental design. (E) A representative image of sparsely labeled MG in the Glast-CreERT2;tdTomato mouse retina. (F) Quantification of the numbers of 1-cell, 2-cell, and larger clones in ten 250 µm hotspot areas from four retinas infected with CCA. (G-I) Eye samples from the time-course EdU analysis (A) were stained for cyclin D1 and Sox2. Representative retinal sections from uninjected eyes (G), mice with EdU injections at D1-2 and harvest at D3 (H), and mice with EdU injections at D11-12 and harvest at D13 (I). Arrowheads point out the EdU+ cells that are negative for cyclin D1 staining. ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer.

We asked why CCA did not drive MG to undergo multiple rounds of cell proliferation. We immunostained the retinal samples harvested for the time-course MG proliferation study (Fig. 2A). Cyclin D1 was undetectable in MG of control retinas (Fig. 2G), but it was robustly overexpressed in the majority of MG within the INL by three days post-CCA injection (Fig. 2H). By a later time point, cyclin D1 expression had disappeared in the EdU+ MG that had migrated to the ONL (Fig. 2I, arrowheads). The other EdU-MG in the ONL, which were likely the cells that have proliferated prior to EdU injections, also ceased to express cyclin D1 (Fig. 2I). In another group of mice injected with CCA on P28, we further confirmed that the levels of cyclin D1 overexpression decreased at 3 weeks post-CCA injection and diminished at 4 months post-CCA injection (Supplementary Fig. 8). In comparison, p27Kip1 suppression by CCA lasted longer, as p27Kip1 level remained low in some MG at 3 weeks and 4 months post-CCA injection (Supplementary Fig. 9-10). These results suggest that cyclin D1 overexpression ceased after the initial cell proliferation, thereby preventing MG from undergoing a second round of proliferation.

scRNA-seq analysis of MG shows suppression of the interferon (IFN) pathway by CCA

To identify changes in gene expression and possible changes in cell fate of MG and their daughter cells, we performed single-cell RNA-Seq (scRNA-Seq) analysis on the mouse retinas that received CCA treatment. To isolate MG in regions of the retina with high viral infection, CCA and a control virus AAV7m8-GFAP-GFP-NT shRNA were 9:1 mixed and co-injected to 4-week-old Glast-CreERT2;tdTomato mice (Fig. 3A). Three weeks post infection, the retina areas with strong GFP expression were collected by dissection, and tdTomato+ MG were isolated by fluorescence-activated cell sorting (FACS) and analysed by scRNA-seq (Fig. 3A). The majority of sorted MG should be infected by CCA. The control group was injected with the AAV7m8-GFAP-GFP-NT shRNA virus only (the same total concentration as the CCA and CCANT groups) to account for any non-specific effect caused by virus injection and/or shRNA expression. As previous studies demonstrated that N-methyl-D-aspartate (NMDA)-induced retinal damage and histone deacetylase inhibitor Trichostatin A (TSA) improve MG proliferation and reprogramming (Hoang et al., 2020; Jorstad et al., 2017; Rueda et al., 2019), a group of CCA-treated mice also received NMDA on day 7 and TSA on day 9 post-CCA injection (CCA+NMDA+TSA, referred to as CCANT) to enhance the reprogramming effect, if any, of CCA.

scRNA-seq analysis of MG at three weeks post-CCA treatment

(A) Schematic illustration of the scRNA-seq experiment. (B) UMAP plot of scRNA-seq data for all MG combined from three groups with control, CCA, or CCANT treatment, with clusters identified based on known marker gene expression. (C) Violin diagram showing expression of retinal cell markers in different cell clusters. (D) Split UMAP plots of control, CCA, and CCANT groups. (E) Proportions of cell clusters within control, CCA, and CCANT groups. (F) Heatmap of top DEGs between cell clusters. Cell clusters are shown in columns, and genes are in rows. Color scale denotes Z score of the normalized gene expression levels. (G) Violin diagram illustrating the expression of IFN pathway genes, MG genes and rod genes across different cell clusters. (H) Feature plots showing normalized gene expression of Stat1, Stat2, Stat3, Gbp6, Irgm1, and Igtp in different cell clusters. CCA, cell cycle activator; CCANT, CCA + NMDA + TSA.

After quality control, 3,758 cells were profiled in the control group, 3,890 cells in the CCA group, and 3,278 cells in the CCANT group (Supplementary Fig. 11-12). Clustering analysis separated the cells into six distinct clusters (Fig. 3B), which are quiescent MG, reactivated MG, MG in G2/M phase, and MG in S phase, rods, and rod-MG, as annotated by the known retinal cell type markers (Fig. 3C). The rods and rod-MG clusters were due to tdTomato leaky expression in native rods and rod mRNA contamination during MG isolation, respectively (Supplementary Fig. 7,13-14). The vast majority of MG (>90%) in the control group are quiescent MG, which express high levels of MG genes such as Glul and Kcnj10 (Fig. 3D-G). In the CCA-treated sample, there was still a small number of proliferating MG in the G2/M or S phase at three weeks after CCA treatment, as shown by cycle state analysis (Fig. 3B-E, Supplementary Fig. 12D,E). ∼70% of MG in the CCA-treated sample formed a separate cluster, which we refer to as reactivated MG (Fig. 3C-E). Reactive gliosis genes (Gfap and Vim) were upregulated while other MG genes (Kcnj10, Glul, Rlbp1, and Aqp4) were downregulated in the reactivated MG compared with the quiescent MG (Supplementary Fig. 15). Interestingly, the top differentially expressed genes (DEGs) between reactivated MG and quiescent MG are the interferon (IFN) pathway genes, including Stat1, Stat2, Gbp6, Irgm1 and Igtp, which were downregulated in the reactivated MG (Fig. 3F-H). IFN signaling is involved in antiviral responses and usually induces cell cycle arrest in infected cells (Durbin et al., 1996; Kaplan et al., 1998). The high level of IFN signaling in the control sample likely resulted from control virus infection, while the pro-proliferative effect of the CCA vector might suppress IFN pathway to promote MG proliferation. In contrast, Stat3, which is often activated by various cytokines or growth factors to promote cell proliferation and survival (Hirano et al., 2000), was not downregulated concurrently with Stat1 and Stat2 (Fig. 3H). Activation of STAT3 signaling in the reactivated MG may also facilitate MG proliferation.

CCA causes a temporary suppression of MG genes, leading to partial dedifferentiation

Despite CCA or CCANT treatment, no neurogenic progenitor cluster or significant upregulation of neurogenic transcription factors such as Ascl1 and Neurog2 were observed (Supplementary Fig. 16), suggesting cell cycle reactivation, even with NMDA and TSA, is insufficient to drive robust neuronal reprogramming. Nonetheless, the scRNA-seq data suggest that CCA treatment led to downregulation of the MG genes, including Glul, Rlbp1, Aqp4, and Kcnj10 (Supplementary Fig. 15). To further verify the change of MG gene expression after CCA treatment, we assessed the level of Glul, the gene encoding Glutamate Synthase, which is highly expressed in the MG, by RNA in situ hybridization on retinal sections (Fig. 4A). Three weeks post-CCA treatment, there was a significant decrease of Glul mRNA level in the MG, regardless their localization in the ONL, OPL or INL (Fig. 4B-D), suggesting a partial repression of the MG gene. By four months post-CCA treatment, the Glul level increased again, suggesting that they eventually retained MG identity and functions (Fig. 4B-D).

Glul mRNA levels decrease in the MG that migrated to the ONL and OPL

(A) Experimental design. (B) In situ hybridization showing Glul mRNA expression. Control retina did not receive any AAV injection. ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer. (C) Magnified views of the highlighted cells in panel (B). (D) Average pixel intensity of Glul mRNA per GFP+ cell. Pixel intensity of MG treated with CCA was normalized to the average pixel intensity of MG in the uninjected eye of the same animal. n=60 GFP+ cells in each retinal layer from three mice, data are presented as mean ± SEM. ****P<0.0001 (unpaired two-tailed Student’s t-test) (D).

We further assessed another MG marker Sox9 in the CCA-treated retinas. Immunostaining confirmed that many MG in the ONL, which have gone interkinetic nuclear migration, decreased or even lost Sox9 expression level at 3 weeks post-CCA treatment (Fig. 5A-E). The suppression of MG gene expression was transient, as most MG had recovered high levels of Sox9 expression by 4 months (Fig. 5E). The results from both assays suggest that MG cell fate was partially suppressed in the MG daughter cells, but it recovered without further stimulus to drive neurogenesis. However, less than 1% of the GFP+ MG in the INL completely lost Sox9 between 3 weeks to 4 months post-CCA treatment (Fig. 5D-E). These Sox9 negative MG exhibited circular nuclear envelop shape and faint GFP signal (Fig. 5D, arrowhead), characteristics of the MG-derived retinal neuron-like cells observed in previous studies (Hoang et al., 2020; Le et al., 2024a, 2024b).

Temporary loss of Sox9 in a subset of MG following CCA treatment

(A) Experimental design. (B) Representative retinal sections of Glast-CreERT2; Sun1:GFP mice without virus injection (control) or mice at three weeks and four months post-CCA injection. Arrowheads highlight Sox9−GFP+ cells. (C-D) Magnified views of the boxed regions in (B). Arrowheads highlight Sox9−GFP+ cells. (E) Quantification of Sox9−GFP+ cells as a percentage of total GFP+ cells in each retinal layer. n=3 mice, data are presented as mean ± SEM. ns=not significant, *P<0.05 (unpaired two-tailed Student’s t-test). ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer.

Rare neurogenesis from MG occurs spontaneously after cell cycle reactivation

To examine whether some MG, although at a very low efficiency, give rise to neurons after CCA treatment, we performed immunostaining for the bipolar marker Otx2. We focused our analysis on EdU+tdT+ MG in the Glast-CreERT2; tdTomato mice to identify de novo neurogenesis from MG (Fig. 6A). Four months post-CCA treatment, we found a small number of MG-derived cells in the INL expressing Otx2 (Fig. 6B-E). Although their proportion was low (∼1% of EdU+tdT+ cells), these were likely the genuine neuron-like cells that have differentiated from MG daughter cells over time, as these cells were not present at an earlier time point (Fig. 6A-E). Rare HuC/D+EdU+tdT+ cells were also found in the lower INL, where the amacrine cells naturally reside, only at four months post-CCA treatment (Supplementary Fig. 17). However, the HuC/D level in the MG-derived cells was lower compared to that of the native amacrine cells (Supplementary Fig. 17). Whether these cells express other bipolar or amacrine cell markers and whether they are functional interneurons that connect with the retinal circuitry needs further investigation. Nonetheless, our findings suggest that cell cycle reactivation alone allows some MG daughter cells to spontaneously become neuron-like cells.

CCA induces de novo genesis of Otx2+ cells from MG

(A) Experimental design. (B) Representative retinal sections of Glast-CreERT2; tdTomato mice without AAV injection (control) or mice at four months post-CCA injection. Sections were co-stained for EdU and Otx2. ONL, outer nuclear layer; OPL, outer plexiform layer; INL, inner nuclear layer; IPL, inner plexiform layer; GCL, ganglion cell layer. (C-D) Magnified views of the highlighted cells in (B). Arrowhead highlights a tdT+EdU+ MG in the OPL that is negative for Otx2 (C), while Arrowhead highlights a tdT+EdU+Otx2+ MG in the INL. (E) Quantification of tdT+EdU+Otx2+ cells as a percentage of total tdT+EdU+ cells. n=3 mice, data are presented as mean ± SEM. ns=not significant, **P<0.01 (one-way ANOVA with Tukey’s post hoc test).

CCA does not cause retinal neoplasia or functional deficit

To assess the long-term effect of CCA treatment, a cohort of C57BL/6J mice were observed for a year following CCA injection (Fig. 7A). Visual function assessments showed no significant differences in visual acuity and electroretinography (ERG) between CCA-treated eyes and uninjected control eyes one year after CCA treatment (Fig. 7B-D). Upon examining all the harvested retinas from the CCA-treated group, no retinal neoplasia was observed. The retinal structure remained intact, with no malignancy or disruptions in the retinal layers or stratification (Fig. 7E-H). These results indicate that CCA treatment did not have a detrimental impact on retinal structure or function in mice, nor did it induce retinal tumours.

CCA does not lead to retinal neoplastic transformation

(A) Experimental design. (B-C) Optomotor and electroretinography (ERG) tests were performed on wild type mice with one eye injected with CCA and the other eye as control (without AAV injection) at one-year post-CCA injection. Visual acuity by the optomotor test (B), b-wave amplitudes of the scotopic ERG under different light intensity in (C), and b-wave amplitudes of the photopic ERG under 30 cd × s/m2 in (D). Data are presented as mean ± SEM. ns=not significant (paired two-tailed Student’s t-test) (B-D). (E-H) Immunostaining for MG marker Sox9. (F,H) Zoomed-in images of the boxed areas in (E) and (G). (I) Quantification of the numbers of Sox9+ cells in control retinas versus CCA-treated retinas at two weeks or one-year post-CCA injection and in age-matched wild type control retinas. Data are presented as mean ± SEM. ns=not significant, ***P<0.001 (one-way ANOVA with Tukey’s post hoc test). (J) Quantification of the number of Sox9+ cells in each retinal layer. Data are presented as mean ± SEM. ns=not significant, **P<0.01 (unpaired two-tailed Student’s t-test).

We used the MG marker Sox9 to examine the MG population in the retina sections harvested at a year post-CCA treatment. In the control retina, Sox9+ MG cells were aligned in the INL (Fig. 7E-F). In the CCA-treated retinas, there was a significant expansion in the number of Sox9+ MG cells, distributed across the ONL, OPL, and INL (Fig. 7G-H). The larger population of Sox9+ MG remained to support the retinal structure and homeostasis, which explains the relatively unaffected retinal structure and function observed on the CCA-treated eyes. We further made comparison of the numbers of Sox9+ cells after one year of CCA treatment and after 2 weeks of CCA treatment. The total numbers of Sox+ cells were similar between these two time points (Fig. 7I), indicating no significant MG loss after MG proliferation. To directly assess whether CCA treatment may lead to the cell death of the MG, especially the MG displaced in the ONL and OPL, we examined the CCA-treated retinas with TUNEL assay and did not observe obvious any TUNEL+ MG at three or six weeks post-CCA injection (Supplement Fig. 18). Since we could only examine cell death at single time points, we cannot completely exclude the possibility that there were small numbers of MG-derived cells dying without being detected. The fact that the numbers of MG in the ONL and OPL did not decrease in the retinas harvested at one-year post-CCA treatment suggests that these MG-derived cells persist without dying or losing the MG identity (Fig. 7J).

Discussion

MG proliferation is a key step towards MG-mediated regeneration in the retina. Previously, it was shown that activation of the canonical Wnt pathway by AAV-mediated overexpression of β-catenin stimulates MG proliferation through the Lin28/let7 pathway (Yao et al., 2016), and cyclin D1 is a direct target gene of both Wnt signaling and Lin28/let7 (Li et al., 2012; Shtutman et al., 1999; Tetsu and McCormick, 1999). Suppression of the Hippo pathway by transgenic expression of YAP5SA induces MG proliferation, accompanied by an increase in cyclin D1 expression (Hamon et al., 2019; Rueda et al., 2019). However, the Hippo and Wnt pathways have downstream genes besides cyclin D1 and serve different cellular functions. It remains unclear whether cyclin D1 activation is necessary or sufficient to drive MG proliferation in these contexts. In this study, we demonstrate that cyclin D1 overexpression alone leads to limited MG proliferation and that the combination of cyclin D1 overexpression and p27Kip1 knockdown is the most potent strategy to drive MG proliferation in mouse retina without injury stimulus.

Our results showed that CCA effectively promoted MG proliferation in both P6 and adult mouse retinas; however, the timing of MG proliferation differed between the two age groups. Following CCA treatment on P6, MG proliferation started as early as day 3-4 post-CCA injection and largely finished by two weeks, peaking around one week after treatment (Fig. 2A). In contrast, following CCA treatment on P28, MG proliferation started later, as not many mitotic MG had migrated to the ONL at one-week post-CCA treatment (Supplementary Fig. 8 and 10). Moreover, proliferating MG were still present three weeks post-treatment, as shown by scRNA-seq analysis (Fig. 3B-E). We speculate that the differences in MG proliferation time are due to variations in AAV transduction efficiencies of retinal cells at different ages. AAV7m8 vectors delivered by intravitreal injections transduce MG more efficiently at P6 before the inner limiting membrane is fully developed. The level of cyclin D1 and p27Kip1 may reach the thresholds permissive for MG cell cycle re-entry faster following CCA treatment on P6 compared to the same treatment in adult mice. This hypothesis is supported by the immunostaining results of both cyclin D1 and p27Kip1 (Fig. 2G-I, supplementary Fig. 8-10). This finding also emphasizes the importance of using high CCA titer (>5 x 1013 genome copies/ml) in adult mice to effectively stimulate MG proliferation.

P27Kip1 and cyclin D1 serve as critical regulators of cell cycle progression, and any abnormalities in their expression may impact cell division, potentially leading to tumor development. Previous studies have shown that mice lacking p27Kip1 are approximately 30% larger and develop pituitary tumours spontaneously (Fero et al., 1996). Cyclin D1 is frequently upregulated in a significant fraction of human cancers, such as breast and respiratory tract tumours (Callender et al., 1994; Zukerberg et al., 1995). The strategy to stimulate MG proliferation for the purpose of retinal regeneration must be carefully weighed against the potential risk of tumorigenesis. In this study, we carefully evaluated the tumorigenic potential of CCA and made the following observations: First, the rate of MG proliferation gradually decreased and ceased over time (Fig. 2A). Second, MG appeared to undergo only a single round of cell division (Fig. 2B-F); and lastly, the retinal structure remained normal without signs of neoplasia even one-year post-treatment (Fig. 7). The absence of continuous MG proliferation may be partly attributed to the non-integrating nature of the recombinant AAV genomes, which become diluted after cell division, leading to reduced transgene expression. Notably, the loss of cyclin D1 overexpression and the recovery of high level p27Kip1 did not occur simultaneously. The former was observed much earlier than the latter (Fig. 2G-I, Supplementary Fig. 8-10). This is also supported by the scRNA-seq data, which showed that Ccnd1 upregulation was lost in the MG three weeks after CCA treatment while Cdkn1b suppression was still evident (Supplementary Fig. 16). These findings suggest that the rapid cessation of cyclin D1 overexpression may be regulated by additional mechanisms. For instance, the anti-proliferative gene Btg2, known to suppress cyclin D1 expression (Wei et al., 2012), was upregulated in the proliferating and reactivated MG (Supplementary Fig. 16). In summary, the risk of tumor formation associated with CCA is minimal.

Interestingly, we observed significant downregulation of interferon (IFN) pathway genes in reactivated MG, a finding consistent with their known anti-proliferative role in immune surveillance and tumor suppression (Durbin et al., 1996; Kaplan et al., 1998). The suppression of IFN pathway may reflect a mechanism by which CCA overcomes intrinsic barriers to proliferation. Our results align with Rueda et al. (2019), who reported that forced YAP5SA expression in NMDA-treated MG similarly downregulated IFN pathway genes in the proliferative MG (Rueda et al., 2019). Given that cyclin D1, a key Hippo pathway target, is upregulated in both paradigms, we propose a shared mechanism of IFN suppression. In addition to regulating cell cycle, cyclin D1 may indirectly control the transcription of Stat1 and Stat2 and/or other IFN pathway genes by interacting with other transcriptional cofactors (Mcmahon et al., 1999; Zwijsen et al., 1998). Notably, Stat3 levels remained stable in the reactivated MG. STAT3 activation, which is primarily induced by cytokines (e.g. IL-6) and growth factors (e.g. EGF), promotes cell proliferation (Hirano et al., 2000), but it inhibits neurogenesis of mouse and avian MG (Jorstad et al., 2020; Todd et al., 2016, 2020). Sustained Stat3 level in our system could explain both the proliferative competence of reactivated MG and their failure to differentiate to neurons. The divergent regulation of Stat1/2 (suppressed) versus Stat3 (sustained) highlights that these pathways are differentially modulated by the cell cycle regulators, warranting further investigation into their different functions in MG reprogramming.

The rods and rod-MG were two unexpected clusters in the scRNA-seq analysis. Upon close examination of the retina of Glast-CreERT2; tdTomato mice without Tamoxifen induction, we found that very few rods (14 tdTomato+ rods in 128 whole retinal sections of eight mouse retina samples screened) were mislabeled by leaky tdTomato expression in a Cre-independent manner (Supplementary Fig. 7). The small rod clusters are likely the native rods mislabeled by tdTomato, appearing in all three groups at similar proportions. The rod-MG cluster, which was characterized by high levels of both MG and rod genes, was located between the rod cluster and the reactivated MG cluster (Fig. 3B-E). This rod-MG cluster was enriched in the CCA and CCANT groups (Fig. 3D,E). Initially, we speculated that this rod-MG cluster might represent MG that had upregulated rod genes. To investigate this, we performed RNA in situ hybridization to assess the levels of Gnat1 and Rho mRNAs in MG freshly isolated from the mouse retina three weeks post-CCA treatment (Supplementary Fig. 14). Despite slight increases of Gnat1 and Rho mRNA signals in the CCA-treated sample compared to the untreated control retina, some MG from the control retina also exhibited some signals of both genes, suggesting that these signals resulted from rod contamination (Supplementary Fig. 14). The enrichment of rod gene expression in CCA-treated MG, as indicated by the scRNA-seq data and RNA in situ hybridization data, may result from a greater number of MG migrating to the ONL and closely contacting with surrounding rods, leading to higher levels of rod contamination. Moreover, we did not observe many cells overexpressing neurogenic genes, such as Ascl1, Neurog2 and Dll3, or rod precursor gene prdm1 (Supplementary Fig. 16). Therefore, we cautiously conclude that CCA alone does not reprogram MG toward a rod cell fate. The lack of MG reprogramming toward rod may be due to the sustained Notch signaling or Stat3 signaling in the reactivated MG (Supplementary Fig. 16).

Following cell division, the majority of the MG retained their glial identity, with around 1% of MG daughter cells expressing markers of bipolar or amacrine cells. This aligns with prior reports showing that a small subset of MG daughter cells expressed the markers of retinal interneurons after cell cycle reactivation either by inhibition of the Hippo pathway or by activation of Wnt signaling (Rueda et al., 2019; Yao et al., 2016). The rarity of neurogenesis raises critical questions about intrinsic heterogeneity within the MG population. We speculate that a small subset of MG may exist in a “primed” state, either expressing elevated levels of neurogenic factors or having a more permissive chromatin state for neurogenic gene expression, which predispose them to differentiate to neurons. However, the inefficiency of neurogenesis driven solely by cell cycle reactivation underscores a key translational barrier. To achieve functional regeneration, future strategies should combine MG cell cycle activation with neurogenic reprogramming factors (e.g. ASCL1) and/or suppression of anti-neurogenic pathways (e.g. Notch, STAT3). Such combinatorial approaches could redirect MG daughter cells from a default glial fate toward neuronal differentiation, offering a viable path for treating retinal degenerative diseases.

Materials and Methods

Animals

Wild type mice (strain C57BL/6J), Rosa26-tdTomato reporter mice (strain B6.Cg-Gt(ROSA)26Sortm14(CAG-tdTomato)Hze/J), Rosa26-Sun1:GFP reporter mice (Strain B6.129-Gt(ROSA)26Sortm5.1(CAG-Sun1/sfGFP)Nat/MmbeJ), and Glast-CreERT2 reporter mice (strain Tg(Slc1a3-cre/ERT)1Nat/J) were purchased from the Jackson Laboratory and were kept on a 12 hour light/12 hour dark cycle. All animal procedures performed were approved by the Hong Kong Department of Health under Animals Ordinance Chapter 340 and City University of Hong Kong animal ethics committee, and animal care was carried out in accordance with institutional guidelines.

AAV plasmids and vectors

The pAAV-GFAP-GFP vector plasmid was cloned by replacing the CMV promoter with a 681bp ABC1D region of the human GFAP gene promoter (Lee et al., 2008) into the pAAV-CMV-GFP vector (Supplementary Fig. 1A). cDNAs encoding mouse cyclin D1 were cloned into AAV plasmids by Gibson ligation. The AAV-shRNA vectors were cloned by replacing the GFP sequence with mCherry-shRNA in the pAAV-GFAP-GFP vector. Three shRNA sequences used in the study were GACTACACAAATCAGCGATTT (non-target shRNA), GCAAGTGGAATTTCGACTTTC (p27 shRNA1), and GCTTGCCCGAGTTCTACTACA (p27 shRNA2). pAAV-GFAP-cyclin D1-p27 shRNA1 and pAAV-GFAP-GFP-NT shRNA were cloned by replacing mCherry sequence with cyclin D1 or GFP from the pAAV-GFAP-mCherry-p27 shRNA1 or pAAV-GFAP-mCherry-NT shRNA, respectively (Supplementary Fig. 2A). The pAAV rep/Cap 2/2, and Adenovirus helper plasmids were obtained from the University of Pennsylvania Vector Core, Philadelphia. The pAAV7m8 plasmid (#64839) was purchased from addgene.

AAV was produced in HEK293T cells (HCL4517; Thermo Scientific) by AAV vector plasmid, rep/cap packaging plasmid, and adenoviral helper plasmid co-transfection followed by iodixanol gradient ultracentrifugation. Purified AAVs were concentrated with Amicon 100K columns (EMD Millipore) to a final titer higher than 5 x 1013 genome copies/ml. Protein gels were run to determine virus titers.

Tamoxifen injection

Intraperitoneal injections of tamoxifen (Sigma) in corn oil were administered to induce the expression of Cre recombinase. Tamoxifen was given at a dosage of 50 mg/kg daily from P23 to P27 to induce reporter expression in the majority of MG, and a single dosage of 15 mg/kg on P20 for sparsely labeling of MG.

AAV injection

Intravitreal injection was performed using a pulled angled glass pipette controlled by a FemtoJet (Eppendorf). The tip of the needle was passed through the sclera at the equator, near the dorsal limbus of the eyeball, and entered the vitreous cavity. The injection volume of AAV (5 x 1013 genome copies/ml) was 0.5μl per eye for P6 injection and 1μl per eye for adult mouse injection.

In situ RNA hybridization

In situ RNA hybridization was performed on retinal sections using the RNAscope Multiplex Fluorescent Detection Reagents V2 kit (Advanced Cell Diagnostics) following the commercial protocols. In brief, retinas were dissected, 4% paraformaldehyde (PFA) fixed, dehydrated in sucrose solution and embedded in optimal cutting temperature (OCT) medium. Retinas were cryosectioned into 20-μm-thick sections and mounted on SuperFrost Plus glass slides (Epredia). The uninjected eye and CCA-injected eye from the same animal were sectioned on the same slide and processed together. After OCT removal with PBS and further dehydration by ethanol, retinal sections were stained with GFP antibody (AB_2307313; Aves Labs) at 4 °C overnight. After three times wash with PBST (PBS with 0.1% Tween-20), retinal sections were hybridized with RNA probes (Advanced Cell Diagnostics Cat.No. 426231-C2 for Mm Glul) at 40 °C for two hours. Following in situ RNA hybridization steps, slides were stained using secondary antibodies (Jackson ImmunoResearch) and DAPI for two hours at room temperature. The fluorescent signals were visualized and captured using Nikon A1HD25 High speed and Large Field of View Confocal Microscope. The mRNA levels were quantified by measuring signal intensity level by ImageJ. Pixel intensity of MG treated with CCA was normalized to the average pixel intensity of MG in the uninjected eye of the same animal.

Immunohistochemistry

Retinas were dissected and fixed in 4% formaldehyde in PBS for 30 min at room temperature and sectioned at 20-μm thickness by cryostat. Retinal sections were blocked in 5% BSA in PBST (PBS with 0.1% Triton X-100), stained with primary antibodies at 4 °C overnight, and washed three times with PBST. Primary antibodies used in this study included rabbit anti-p27kip1(1:200, PA5-16717; Thermo Fisher); rabbit anti-cyclin D1 antibody (1:300, 26939-1-AP; Proteintech), goat anti-Otx2 antibody (1:200, AF 1979; R&D systems), mouse anti-HuC/D antibody (1:200, A21271; Thermo Fisher), rabbit anti-GFAP antibody (1:500, Z0334, DAKO), goat anti-Sox2 antibody (1:500, AF2018, R&D systems), and rabbit anti-Sox9 antibody (1:1000, AB5535; Millipore). Sections were stained using secondary antibodies (Jackson ImmunoResearch) for two hours at room temperature. Cell nuclei were counterstained with DAPI (Sigma). TUNEL was performed using an TUNEL BrightRed Apoptosis Detection Kit (Vazyme) according to the manufacturer’s protocol.

EdU incorporation and BrdU detection assay

5’-ethynyl-2’-deoxyuridine (EdU, 50 mg/kg, Abcam ab146186) or 5-bromo-2’ deoxyuridine (BrdU,100 mg/kg, Abcam 142567) was injected intraperitoneally to label the cells in the S phase. EdU staining was performed using the Click-iT™ EdU Alexa Fluor™ Imaging Kit (C10337). For BrdU detection, the retinal sections were incubated with 2 M HCl for 1 hour at room temperature. The sections were rinsed with PBST and incubated with a blocking buffer containing 5% BSA in PBST for 2 hours at room temperature. Primary antibody mouse anti-BrdU (1:300, ab8152; Abcam) was stained overnight at 4°C, and secondary antibodies (Jackson ImmunoResearch) were stained for 2 hours at room temperature.

Single-cell RNA-seq

Library preparation and sequencing of single cells

The FACS-sorted cells from each sample were filtered through a 40-µm strainer (pluriStrainer) and processed with Chromium Next GEM Single Cell 3′ GEM, Library & Gel Bead Kit v3.1 (1000269), Chromium Next GEM Chip G Single Cell Kit (1000127) and Chromium Controller (10x Genomics) according to the manufacturer’s protocol. The constructed libraries were sent to Novogene (Beijing) for NovaSeq paired-end 150 bp sequencing and produced 330 Gb of raw data.

Preprocessing, filtering and clustering of scRNA data

Sequencing results were preprocessed with 10x Genomics Cell Ranger 6.1.1. (Zheng et al., 2017) for demultiplexing, barcode assignment, and unique molecular identifier (UMI) quantification using a standard pipeline for 10X for each of the three samples. We used indexed mouse genome mm10 with added tdTomato and GFP transgenes as a reference. Reads mapping to introns covered more than 34% of all reads. Therefore, they were included in feature counts using ‘include_introns’ option in cellranger count step of the pipeline. Cell Ranger aggregation was applied for merging runs from three samples.

Quality control, filtering, dimensional reduction, and clustering of the data were carried out using the Seurat package in R (Stuart et al., 2019). The cells with less than 2,000 expressed features, total UMI count higher than 100,000 and percentage of mitochondrial transcripts more than 15% were disregarded. Features of the Y chromosome, as well as Xist and Tsix genes of the X chromosome, were excluded to avoid confusing effects of gender.

Further filtering of cells was carried out by only including cells expressing tdTomato (Supplementary Fig. 11). Potential cell duplicates were estimated using DoubletFinder method (McGinnis et al., 2019) implemented in R (with parameter settings PC=10, pK=0.28, pN=0.3). The threshold for the expected doublet formation rate was set to 10% to reflect the assumption that cell duplicates can appear relatively commonly in the MG cells. Detected duplicates were excluded from the data (Supplementary Fig. 11G,H).

For the remaining cells, Uniform Manifold Approximation and Projection (UMAP) dimension reduction based on eight principal components (PCs) was applied, and cells were clustered using the graphical clustering method in Seurat. Cell types were identified using known marker genes (Clark et al., 2019; Hoang et al., 2020). After removing doublets, cell types in clusters 4-9 other than MG, Rod-MG, and Rod included only a few cells, and 10-30% of the cells were from the control sample, suggesting that these cells are likely to be contaminated (Supplementary Fig. 11I,J). After excluding cells in clusters 4-9, the new UMAP and clustering (Supplementary Fig. 12A) revealed two subclusters of Rod-MG cells (Clusters C3-C4, Supplementary Fig. 12A). The cells in the subcluster C4 exhibited similar proportional contributions from all three groups and expressed lower level of rod-specific genes such as Rho (Supplementary Fig. 12B,C). These cells were again defined as contaminated and excluded from the data. The final UMAP was constructed using seven PCs, and the same PCs were used for clustering of cells (Fig. 3B).

Analysis of scRNA-seq data

The cell cycle state of each cell was defined by the Cell Cycle Scoring function in the Seurat R package. The cells with G2/M and S scores lower than 0.1 were defined to be proliferating (Supplementary Fig. 12D,E). Differential gene expression analysis between different cell types or three treatment groups was performed using Deseq2 (Supplementary table 1) (Love et al., 2014). Enrichment analysis of top 200 up- and downregulated genes was performed for Gene Ontologies (GO), KEGG and Reactome pathways using Gprofiler2 R interface to g:Profiler (Raudvere et al., 2019).

The optomotor and electroretinography tests

Mouse visual acuity was measured using an Optometry System (Cerebral Mechanics Inc.) following the published protocol (Douglas et al., 2005). Testing was done with a grating of 12 degrees/second drifting speed and 100% contrast. The injected right eyes (CCA-treated) and uninjected left eyes (Ctrl) were tested independently for counterclockwise and clockwise grating rotations, respectively. A staircase procedure was used, in which the observer tested low to high visual acuity. Each animal was tested for about 10-15 minutes per session.

ERG was measured using an Espion E3 System (Diagonsys LLC Inc.) as previously described (Hoang et al., 2023). Animals were dark-adapted overnight before the test. After inducing anesthesia with a ketamine: xylazaine injection intraperitoneally, the pupils of both eyes were dilated using Mydrin®-P Ophthalmic Solution. Once the pupils were fully dilated, the eyes were maintained moist using a topical gel. The measurement electrodes were then applied to the cornea of both eyes, while the ground electrodes were attached to the mouth and tail. All these steps were performed in the darkroom under dim red light. For scotopic ERG recordings, a multiple 530nm light with different intensities (increments from 0.01 cd.s/m2 to 30 cd.s/m2) were elicited to stimulate scotopic responses in a specific time interval. For photopic ERG recordings, 5 mins exposure under 10 cd.s/m2 light intensity was adopted to inhibit the rod function. The photopic response was measured by multiple flashes of 30 cd.s/m2 intensity in the illuminated background (10 cd.s/m2). The average amplitude and implicit time of a- and b-wave were recorded and exported for further analysis. The b-wave amplitude of the step-wise scotopic responses and that of photopic response at 30 cd.s/m2 were shown in figure.

Statistics

Data were presented as mean ± SEM in all figures. Sample sizes and statistical analysis were indicated for each experiment in figure legend. ANOVA with Tukey’s test was performed to compare multiple groups, and Student’s t-test to compare two groups. A P value < 0.05 was considered statistically significant. GraphPad Prism was used to perform statistical analysis and make figures.